---
title: "Bacterial Stress Responses Lower mRNA-Protein Level Correlations"
author: "Sena Gizem Suer, Yi Yang Lim, Rabia Şen, Cemal Erdem, André Mateus and Kemal Avican"
date: "2026-02-18"
output:
  rmdformats::downcute:
    code_folding: hide
    self_contained: true
    thumbnails: false
    lightbox: trues
pkgdown:
  as_is: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.height = 8,
  fig.width = 16
)
```

```{r load-environment, results='hide'}
# Load library
library(fs)
library(purrr)
library(readxl)
library(patchwork)
library(stringr)
library(tidyverse)
library(limma)

# Load sources functions
source("sources/utils.R")
```

```{r configuration, results='hide', error=TRUE}
#' Species Abbreviation Mapping
#' Maps full species names to abbreviations for data parsing
SPECIES <- c(
  "Salmonella_enterica",
  "Staphylococcus_aureus",
  "Yersinia_pseudotuberculosis"
)

SPECIES_ABBV_MAP <- c(
  "Salmonella_enterica" = "SALMT",
  "Staphylococcus_aureus" = "MSSA",
  "Yersinia_pseudotuberculosis" = "YPSTB"
)

#' Treatment List

STRESS_TREATMENTS <- c("As", "Bs", "Li", "Nd", "Ns", "Oss", "Oxs", "Sp", "Tm")
STRESS_TREATMENTS_RNA <- c(STRESS_TREATMENTS, "Mig")
STRESS_TREATMENTS_PROT <- c(STRESS_TREATMENTS, "Hyp")

#' Control and stress treatment conditions
TREATMENTS_LIST_RNA <- c("Ctrl", STRESS_TREATMENTS_RNA)

PLOT_SPECIES_NAMES <- c(
  "Salmonella_enterica" = "S. enterica Typhimurium",
  "Staphylococcus_aureus" = "S. aureus",
  "Yersinia_pseudotuberculosis" = "Y. pseudotuberculosis"
)


#' Exposure Time Label Mapping
#' Maps numeric exposure times (minutes) to display labels
EXPOSURE_TIME_LABELS <- c(
  "10" = "10",
  "20" = "20 min",
  "30" = "30 min",
  "240" = "4 hours",
  "241" = "\u22654 hours"
)

#' Measurement Display Labels
MEASUREMENT_LABELS <- c(
  "corr_intensity" = "Intensity",
  "corr_RI" = "RI",
  "corr_iBAQ" = "iBAQ",
  "corr_iBAQ_mc" = "iBAQ MC"
)

#' Treatment Shape Mapping (ggstar)
TREATMENT_SHAPES <- c(
  "As" = 1,
  "Bs" = 14,
  "Ctrl" = 15,
  "Hyp" = 11,
  "Li" = 7,
  "Nd" = 8,
  "Ns" = 4,
  "Oss" = 22,
  "Oxs" = 21,
  "Sp" = 23,
  "Tm" = 2
)

# Uncomment the quantification you want to use
# PROTEIN_QUANTIFICATION_METHOD <- "RI" # Intensity divided by the number of razor peptide
# PROTEIN_QUANTIFICATION_METHOD <- "Intensity" # Direct intensity from TMT analyses
# TODO boxplot with protein lenght
PROTEIN_QUANTIFICATION_METHOD = "iBAQ" # Total intensity divided by the number of expected peptide after trypsin digestion
# PROTEIN_QUANTIFICATION_METHOD = "iBAQ_MC" Same but miss cleavage parameter filter added

# --- Directories ---------------------------------------------------------------
dirs <- list(
  DATA = file.path("DATA"),
  RNA_DATA = file.path("DATA", "Read_counts"),
  PROTEIN_DATA = file.path("DATA", "Proteome"),
  PROTEIN_FASTA = file.path("DATA", "Fasta_files"), # Uncomment for iBAQ quantification
  ESSENTIAL_GENES = file.path("DATA", "Essential genes"),
  ENRICHMENT = file.path("DATA", "Enrichment")
)

walk(dirs, check_dir, return_error = TRUE)

OUTPUT_PATH <- file.path("Analyses_iBAQ")
FIGURE_FORMAT <- c("png", "svg", "pdf") # "eps", "ps", "tex" (pictex), "jpeg", "tiff", "png", "bmp" or "wmf" (windows only)
MSG_FIG_FORMAT <- paste0(".", paste(FIGURE_FORMAT, collapse = "/."))
dir.create(OUTPUT_PATH)
for (species in SPECIES) dir.create(file.path(OUTPUT_PATH, species))

# --- Required files ------------------------------------------------------------
files <- list(
  SAMPLE_MAPPING = file.path(dirs$PROTEIN_DATA, "sample_mapping.xlsx")
)

walk(files, ~if (!file_exists(.x)) stop(.x, " not found"))

# --- Expose to environment (equivalent to your original variables) -------------
list2env(c(dirs, files), envir = .GlobalEnv)
```

# Data loading and processing
```{r load-tpm, results='hide', message=TRUE}
source("sources/read_data.R")
# Load RNA tpm data
rna_files <- list.files(RNA_DATA, pattern = "\\.xlsx$", full.names = TRUE)

tpm_data <- map_dfr(rna_files, function(rna_file) {
  species <- get_species_tpm(rna_file)
  abbv <- SPECIES_ABBV_MAP[species]
  read_tpm(rna_file, abbv) %>%
    mutate(
      Treatment = ifelse(Treatment == "Mig", "Hyp", Treatment),
      Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE)
    ) %>%
    filter(!is.na(Protein_id), !is.na(TPM)) %>%
    mutate(Species = species)
})
```

```{r load-prot, results='hide'}
# source("sources/read_data.R")
# Load proteomic data
sample_mapping <- read_excel(SAMPLE_MAPPING, col_names = TRUE) %>%
  mutate(Replicate = paste0("R", Replicate))
protein_files <- list.files(
  PROTEIN_DATA,
  pattern = "\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)

filter_log <- list()

filtered_data <- map_dfr(protein_files, function(protein_file) {
  species <- sub(".*/(\\w+)_R\\d+\\.tsv$", "\\1", protein_file)
  replicate <- sub(".*/\\w+_(R\\d+)\\.tsv$", "\\1", protein_file)

  tmt_raw <- read_tmt(protein_file)
  filtered_raw <- filter_tmt(tmt_raw, species, replicate)

  filter_log[[species]][[replicate]] <<- filtered_raw$filter_log

  filtered_raw$filtered_tmt %>%
    mutate(Species = species, Replicate = replicate)
}) %>%
  left_join(sample_mapping, by = c("Species", "Replicate"), relationship = "many-to-many")

# TODO put this in a function
# Build and write filter log
filter_log_df <- map_dfr(filter_log, ~bind_rows(.x)) %>%
  mutate(across(-c("species", "replicate"), as.integer))

message("TMT filtering summary:\n")

write.table(
  filter_log_df,
  file.path(OUTPUT_PATH, "filter_log.tsv"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

# --- Filter for complete proteins ---
completeness_result <- filter_complete_proteins(filtered_data)
tmt_data <- completeness_result$filtered_data
removed_proteins <- completeness_result$removed_proteins

# Write removed proteins table for manual review
if (nrow(removed_proteins) > 0) {
  write.table(
    removed_proteins %>% arrange(Species, Protein_id),
    file.path(OUTPUT_PATH, "removed_proteins_incomplete.tsv"),
    sep = "\t", row.names = FALSE, quote = FALSE
  )
  message("Removed proteins written to: removed_proteins_incomplete.tsv")
}
```

# Trancriptomic and Proteomic analysis of 3 human pathogens under 10 relevant stress conditions

```{r load-annotation, results='hide'}
annotation_files <- list.files(
  PROTEIN_DATA,
  pattern = "^annotation\\.xlsx$",
  recursive = TRUE,
  full.names = TRUE
)

annotation_data <- map_dfr(annotation_files, function(annotation_file) {
  species <- path_file(path_dir(annotation_file))
  read_excel(
    annotation_file,
    col_names = TRUE, na = "N/A"
  ) %>%
    select(locus_tag, old_locus_tag) %>%
    mutate(
      Protein_id = if_else(
        is.na(locus_tag),
        old_locus_tag,
        locus_tag
      ),
      Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE),
      Species = species
    )
})
```

## Percentages of CDSs detected by transcriptomics and proteomics for the three species
```{r cds-distribution}
source("sources/cds_distribution.R")

detected_cds_df <- get_detected_cds(annotation_data, tpm_data, tmt_data)
detected_cds_file <- file.path(OUTPUT_PATH, "Detected_CDS.tsv")
print(detected_cds_df)
write.table(detected_cds_df, detected_cds_file, sep = "\t")
p <- plot_barplot_detected_cds(
  detected_cds_df,
  position = "overlay"
)

print(p)

output_file <- file.path(OUTPUT_PATH, "CDS_detected")
save_plot(
  plot = p,
  filepath = output_file,
  width = 8,
  height = 12,
)
message(
  "CDS detected for mRNA and prtoein saved: ", output_file, MSG_FIG_FORMAT
)
```

## Distribution of transcript abundance across all conditions in the three species

- All condition: light blue,
- Fraction of proteins detected: yellow,
- Transcripts with no corresponding protein detected: dark blue.

```{r plot-transcript-abundance}
# Shared x_max across all species for visual consistency
source("sources/cds_distribution.R")
tpm_log2_data <- tpm_data %>%
  group_by(Species) %>%
  log2_transform(
    value_col = "TPM",
    group_cols = c("Species", "Protein_id", "Treatment"),
    pseudocount = TRUE
  ) %>%
  filter(is.finite(mean_log2)) %>%
  rename(mean_log2_TPM = mean_log2)

x_max <- tpm_log2_data %>%
  summarise(x_max = ceiling(max(mean_log2_TPM, na.rm = TRUE) / 2.5) * 2.5) %>%
  pull(x_max)

species_list <- unique(tpm_log2_data$Species)

plots <- imap(species_list, function(species, i) {
  message("Plotting TPM detection for: ", species)

  classified <- classify_tpm_groups(tpm_log2_data, tmt_data, species)

  p <- plot_tpm_detection_species(classified, species, x_max = x_max)

  output_sp <- file.path(OUTPUT_PATH, species, "TPM_detection_distribution")
  save_plot(
    plot = p,
    filepath = output_sp,
    width = 16,
    height = 5
  )

  message(
    "CDS distribution for ", species, " saved: ", output_sp, MSG_FIG_FORMAT
  )
  if (i > 1) p <- p + theme(axis.title.y = element_blank())
  p
})

# Combined figure — all species side by side
combined <- wrap_plots(plots, nrow = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

print(combined)

output_file <- file.path(OUTPUT_PATH, "TPM_detection_distribution")
save_plot(
  plot = combined,
  filepath = file.path(OUTPUT_PATH, "TPM_detection_distribution"),
  width = 16,
  height = 4
)

message("Combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

# mRNA and protein level correlation under stress

## Protein quantification
```{r in-silico-digest, eval = PROTEIN_QUANTIFICATION_METHOD %in% c("iBAQ", "iBAQ_MC")}
# This chunk only runs if PROTEIN_QUANTIFICATION_METHOD is "iBAQ" or "iBAQ_MC"
# For iBAQ quantification
MIN_PEPTIDE_LEN <- 6
MAX_PEPTIDE_LEN <- 30
IBAQ_MISS_CLEAVAGE <- 2
source("sources/protein_quantification.R")

fasta_files <- list.files(PROTEIN_FASTA, pattern = "\\.fa$", full.names = TRUE)
results <- process_tryptic(
  fasta_paths = fasta_files,
  output_path = OUTPUT_PATH,
  min_peptide_len = MIN_PEPTIDE_LEN,
  max_peptide_len = MAX_PEPTIDE_LEN,
  missed_cleavages = ifelse(PROTEIN_QUANTIFICATION_METHOD == "iBAQ", 0, IBAQ_MISS_CLEAVAGE)
)
protein2tryptic <- results$observable_peptides
all_raw_digests <- results$all_raw_digests
```
```{r protein-quantification}
source("sources/protein_quantification.R")

# Protein quantification method — uncomment the desired one:
# PROTEIN_QUANTIFICATION_METHOD <- "RI"       # Intensity divided by number of razor peptides
# PROTEIN_QUANTIFICATION_METHOD <- "Intensity" # Direct intensity from TMT analyses
# PROTEIN_QUANTIFICATION_METHOD <- "iBAQ"        # Total intensity divided by number of expected tryptic peptides
# PROTEIN_QUANTIFICATION_METHOD <- "iBAQ_MC"  # Same as iBAQ but with missed cleavage filter

message("Quantification method: ", PROTEIN_QUANTIFICATION_METHOD)

# Dispatch table: method -> quantification function
quant_fn <- switch(PROTEIN_QUANTIFICATION_METHOD,
                   RI = function(x) RI_quantification(x),
                   Intensity = function(x) Intensity_quantification(x),
                   iBAQ = function(x) ibaq_quantification(x, protein2tryptic, "n_peptides", OUTPUT_PATH),
                   iBAQ_MC = function(x) ibaq_quantification(x, protein2tryptic, "n_peptides", OUTPUT_PATH),
                   stop("Unknown quantification method: ", PROTEIN_QUANTIFICATION_METHOD)
)

quant_data <- quant_fn(tmt_data)
completeness_result <- filter_complete_proteins(quant_data)
protQ_data <- completeness_result$filtered_data
removed_proteins <- completeness_result$removed_proteins

if (nrow(removed_proteins) > 0) {
  removed_path <- file.path(
    OUTPUT_PATH,
    paste0(PROTEIN_QUANTIFICATION_METHOD, "_removed_proteins_incomplete.tsv")
  )
  write.table(
    removed_proteins %>% arrange(Species, Protein_id),
    removed_path,
    sep = "\t", row.names = FALSE, quote = FALSE
  )
  message(PROTEIN_QUANTIFICATION_METHOD, " removed proteins written to: ", removed_path)
}
protQ_path <- file.path(
  OUTPUT_PATH,
  paste0(PROTEIN_QUANTIFICATION_METHOD, "_data.tsv")
)
write.table(protQ_data, protQ_path, sep = "\t", row.names = FALSE, quote = FALSE)
message(
  PROTEIN_QUANTIFICATION_METHOD,
  " protein quantification data saved: ",
  protQ_path
)
```

## mRNA-protein levels correlation under different stress conditions and control condition

```{r rna-protein-correlation, fig.height = 24}
source("sources/correlation.R")
protQ_meanlog2 <- paste0(PROTEIN_QUANTIFICATION_METHOD, "_meanlog2")
common_ids <- extract_common_ids(tpm_log2_data, protQ_data)

tpm_log2_data <- tpm_log2_data %>% filter(Protein_id %in% common_ids)
protQ_data <- protQ_data %>% filter(Protein_id %in% common_ids)
corr_rna_prot <- calculate_correlation(tpm_log2_data, protQ_data, protQ_meanlog2)
p <- plot_correlation(
  corr_rna_prot, protQ_meanlog2, PROTEIN_QUANTIFICATION_METHOD, OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0("TMP_", PROTEIN_QUANTIFICATION_METHOD, "_correlation_panels")
)
save_plot(
  p,
  output_file,
  width = 16,
  height = 20,
  units = "in"
)
message("Combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

# Essential genes have higher mRNA-protein level correlations in the three species

```{r get-essential-data}
source("sources/essential_analyses.R")
log2_mean_sd_transform <- function(data, value_col, group_cols = c("Species", "Protein_id"), pseudocount = FALSE) {
  mean_col <- paste0("mean_log2_", value_col)
  sd_col <- paste0("sd_log2_", value_col)

  mean_data <- data %>%
    log2_transform(value_col, group_cols = group_cols, pseudocount = pseudocount) %>%
    filter(is.finite(mean_log2)) %>%
    dplyr::rename(!!mean_col := mean_log2)

  sd_data <- data %>%
    log2_mean_sd(value_col, sd_col, group_cols = group_cols, pseudocount = pseudocount)

  left_join(mean_data, sd_data, by = group_cols)
}

essential_files <- list.files(ESSENTIAL_GENES, pattern = "\\.xlsx$", full.names = TRUE)

essential_df <- map_dfr(essential_files, function(file) {
  species <- str_extract(basename(file), "Essential_genes_([A-Za-z]+)", group = 1)
  tibble(
    Species = species_essential2tpm_map[species],
    Protein_id = read_essential_vec(file)
  )
})

tpm_glob_log2mean_data <- tpm_data %>%
  filter(Protein_id %in% common_ids) %>%
  log2_mean_sd_transform("TPM", pseudocount = TRUE)

protQ_glob_log2mean_data <- protQ_data %>%
  filter(Protein_id %in% common_ids) %>%
  log2_mean_sd_transform(PROTEIN_QUANTIFICATION_METHOD, pseudocount = FALSE)

essential_counts_table <- essential_df %>%
  group_by(Species) %>%
  summarise(n_essential_total = n_distinct(Protein_id), .groups = "drop") %>%
  left_join(
    essential_df %>%
      semi_join(tpm_glob_log2mean_data, by = c("Species", "Protein_id")) %>%
      group_by(Species) %>%
      summarise(n_essential_rna = n_distinct(Protein_id), .groups = "drop"),
    by = "Species"
  ) %>%
  left_join(
    essential_df %>%
      semi_join(protQ_glob_log2mean_data, by = c("Species", "Protein_id")) %>%
      group_by(Species) %>%
      summarise(n_essential_prot = n_distinct(Protein_id), .groups = "drop"),
    by = "Species"
  )

print(essential_counts_table)
```

## Mean mRNA expression frequency of essential and non-essential genes

```{r essential-distribution-rna}
source("sources/essential_analyses.R")
p <- plot_essential_distribution(
  all_df = tpm_glob_log2mean_data,
  essential_df = essential_df,
  log2_col = "mean_log2_TPM",
  x_lab = "TPM",
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0("TPM", "_essential_distribution")
)
save_plot(
  p,
  output_file,
  width = 16,
  height = 5,
  units = "in"
)
message("TPM essential distribution combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

## Mean protein expression frequency of essential and non-essential genes
```{r essential-distribution-prot}
source("sources/essential_analyses.R")
p <- plot_essential_distribution(
  all_df = protQ_glob_log2mean_data,
  essential_df = essential_df,
  log2_col = paste0("mean_log2_", PROTEIN_QUANTIFICATION_METHOD),
  x_lab = PROTEIN_QUANTIFICATION_METHOD,
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0(
    PROTEIN_QUANTIFICATION_METHOD,
    "_essential_distribution"
  )
)
save_plot(
  p,
  output_file,
  width = 16,
  height = 5,
  units = "in"
)
message(
  PROTEIN_QUANTIFICATION_METHOD,
  " essential distribution combined figure saved: ",
  output_file,
  MSG_FIG_FORMAT
)
```

## Essential genes and all genes mRNA-protein level correlation

mRNA-protein level correlations were measured with Pearson correlation coefficient (R).
p values were calculated using a Wilcoxon rank sum test.

```{r essential-correlation}
source("sources/essential_analyses.R")
essential_tpm_df <- tpm_log2_data %>%
  semi_join(essential_df, by = c("Species", "Protein_id")) %>%
  filter(is.finite(mean_log2_TPM))
essential_protQ_df <- protQ_data %>%
  semi_join(essential_df, by = c("Species", "Protein_id")) %>%
  filter(is.finite(.data[[protQ_meanlog2]]))

corr_essential_rna_prot <- calculate_correlation(
  essential_tpm_df, essential_protQ_df, protQ_meanlog2
)

p <- plot_essential_vs_all_correlation(
  corr_rna_prot,
  corr_essential_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD,
  OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0(
    PROTEIN_QUANTIFICATION_METHOD,
    "_essential_vs_all_correlation"
  )
)
save_plot(
  p,
  output_file,
  width = 16,
  height = 5,
  units = "in"
)
message(
  "mRNA-Protein correlation all genes vs essential genes combined figure saved: ",
  output_file,
  MSG_FIG_FORMAT
)
```

## Frequency of standard deviation of mRNA expression
```{r essential-sd-rna}
source("sources/essential_analyses.R")
p <- plot_sd_distribution(
  all_df = tpm_glob_log2mean_data,
  essential_df = essential_df,
  sd_col = "sd_log2_TPM",
  x_lab = "TPM",
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0("TPM", "_essential_sd_distribution")
)
save_plot(
  p,
  output_file,
  width = 16,
  height = 5,
  units = "in"
)
message("TPM essential standard deviation distribution combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

## Frequency of standard deviation of protein expression
```{r essential-sd-prot}
source("sources/utils.R")
source("sources/essential_analyses.R")
p <- plot_sd_distribution(
  all_df = protQ_glob_log2mean_data,
  essential_df = essential_df,
  sd_col = paste0("sd_log2_", PROTEIN_QUANTIFICATION_METHOD),
  x_lab = PROTEIN_QUANTIFICATION_METHOD,
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0(
    PROTEIN_QUANTIFICATION_METHOD,
    "_essential_sd_distribution"
  )
)
save_plot(
  p,
  output_file,
  width = 16,
  height = 5,
  units = "in"
)
message(
  PROTEIN_QUANTIFICATION_METHOD,
  " essential standard deviation distribution combined figure saved: ",
  output_file,
  MSG_FIG_FORMAT
)
```

# Stress responses differentially affect mRNA-protein correlation

## mRNA-protein level correlations for conditions exposed stressors

```{r exposure-time}
source("sources/correlation.R")
treatment_to_exposure_time <- c(
  "As" = 10,
  "Bs" = 10,
  "Hyp" = 240,
  "Li" = 10,
  "Nd" = 30,
  "Ns" = 10,
  "Oss" = 10,
  "Oxs" = 10,
  "Sp" = 241, # Not actual time, but used as value >240 -> >4H
  "Tm" = 20
)

print(corr_rna_prot)
exposure_corr_df <- corr_rna_prot %>%
  filter(Treatment != "Ctrl") %>%
  distinct(Species, Treatment, R) %>%
  mutate(
    exposure_time = treatment_to_exposure_time[Treatment],
    exposure_label = factor(
      EXPOSURE_TIME_LABELS[as.character(exposure_time)],
      levels = EXPOSURE_TIME_LABELS
    )
  )
print(exposure_corr_df)
stats_df <- compute_exposure_correlations(exposure_corr_df)
print(stats_df)
label_pos <- build_label_position(exposure_corr_df, stats_df)

p <- build_exposure_plot(
  df = exposure_corr_df,
  label_pos = label_pos
)

print(p)
output_file <- file.path(OUTPUT_PATH, paste0("correlation_vs_exposure_time"))
save_plot(
  p,
  output_file,
  width = 12,
  height = 5,
  units = "in"
)
message("Correlation vs exposure time saved : ", output_file, MSG_FIG_FORMAT)
```

## Protein normalization

```{r normalization}
source("sources/normalization.R")

NORMALIZATION_METHOD <- "VSN" # Go to supplementary to run normalization comparison

# Single method normalization — returns tidy tibble
norm_data <- run_normalization(
  protQ_data,
  protQ = PROTEIN_QUANTIFICATION_METHOD,
  method = NORMALIZATION_METHOD,
  output_path = OUTPUT_PATH
)
```

## Stress responses differentially affect mRNA-protein correlation

```{r read-DEG}
DEG_PVAL_THRESHOLD <- 0.05
DE_FC_THRESHOLD <- 0.5
source("sources/read_data.R")
deg_raw <- map_dfr(rna_files, function(filepath) {
  species <- tools::file_path_sans_ext(
    str_extract(basename(filepath), "^[A-Za-z]+_[A-Za-z]+")
  )
  message("  Loading DEG data for: ", species)
  deg_data_sp <- read_deg_data(filepath) %>%
    mutate(Species = species) %>%
    dplyr::select(Species, Protein_id, Treatment, log2FC, fdr, max_mean)
}) %>%
  filter(!is.na(log2FC), !is.na(fdr)) %>%
  filter(Protein_id %in% common_ids)

print

deg_data <- deg_raw %>%
  group_by(Species, Treatment) %>%
  summarise(
    n_de = sum(fdr < DEG_PVAL_THRESHOLD & abs(log2FC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = n(),
    pct_de = 100 * n_de / n_total,
    de_ids = list(Protein_id[fdr < DEG_PVAL_THRESHOLD &
                               abs(log2FC) > DE_FC_THRESHOLD &
                               !is.na(fdr)]),
    .groups = "drop"
  )
```

```{r compute-DEP}
source("sources/differential_expression.R")
DEP_PVAL_THRESHOLD <- 0.05

species_list <- unique(norm_data$Species)

dep_data <- map_dfr(species_list, function(species) {
  message("  Running limma DEP for: ", species)

  sp_data <- norm_data %>%
    filter(
      Species == species,
      Treatment %in% c("Ctrl", STRESS_TREATMENTS_PROT),
      Protein_id %in% common_ids, # add this
      is.finite(value)
    )

  prot_expr <- build_expression_matrix(sp_data)

  run_limma_dep(
    mat = prot_expr$mat,
    metadata = prot_expr$metadata,
    species = species,
    pval_threshold = DEP_PVAL_THRESHOLD,
    fc_threshold = DE_FC_THRESHOLD
  )
})
```

```{r de-comparison}
source("sources/differential_expression.R")
plots <- plot_de_barplot_all(deg_data, dep_data, OUTPUT_PATH)
combined <- wrap_plots(plots, nrow = 1) +
  plot_layout(guides = "collect", axes = "collect_y") &
  theme(legend.position = "top")

print(combined)

output_file <- file.path(OUTPUT_PATH, "DEG_DEP_barplot")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 16, height = 5
)
message("  Combined plot saved: ", output_file, MSG_FIG_FORMAT)
```


```{r de-correlation}
source("sources/differential_expression.R")
plots <- plot_de_vs_correlation_all(
  deg_data,
  dep_data,
  corr_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD,
  OUTPUT_PATH
)
combined <- wrap_plots(plots, nrow = 2) +
  plot_layout(guides = "collect", axes = "collect_x") &
  theme(legend.position = "right")

print(combined)
output_file <- file.path(OUTPUT_PATH, "DE_vs_correlation")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 16, height = 10
)
```

# Identification of genes whose mRNA-protein levels are discordantly regulated under stress conditions

```{r discordant-data-generation}
source("sources/discordant_analyses.R")
dep_raw <- map_dfr(species_list, function(species) {
  sp_data <- norm_data %>%
    filter(
      Species == species,
      Treatment %in% c("Ctrl", STRESS_TREATMENTS_PROT),
      Protein_id %in% common_ids,
      is.finite(value)
    )
  prot_expr <- build_expression_matrix(sp_data)
  extract_dep_raw(prot_expr$mat, prot_expr$metadata, species)
})

discordance_df <- build_discordance_clusters(
  deg_raw, dep_raw,
  fc_threshold = DE_FC_THRESHOLD,
  pval_threshold = DEP_PVAL_THRESHOLD
)

# Save
write.table(discordance_df,
            file.path(OUTPUT_PATH, "discordance_clusters.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE
)

# Summary with discordant and non-discordant counts
all_proteins <- inner_join(
  deg_raw %>% distinct(Species, Protein_id),
  dep_raw %>% distinct(Species, Protein_id),
  by = c("Species", "Protein_id")
)

discordant_proteins <- discordance_df %>%
  distinct(Species, Protein_id, cluster)

summary_table <- discordant_proteins %>%
  count(Species, cluster) %>%
  arrange(Species, cluster) %>%
  bind_rows(
    all_proteins %>%
      anti_join(discordant_proteins, by = c("Species", "Protein_id")) %>%
      count(Species) %>%
      mutate(cluster = "Non-discordant")
  ) %>%
  arrange(Species, cluster) %>%
  group_by(Species) %>%
  mutate(
    total = sum(n),
    pct = round(100 * n / total, 1)
  ) %>%
  ungroup()

write.table(summary_table, sep = "\t", row.names = FALSE, quote = FALSE)
```

## fold changes of mRNA and protein levels for each of the 10 stress conditions in comparison to control for the three species.

The 6 gene clusters whose mRNA and protein level have different discordant regulation patterns are shown in different colors.

```{r discordant-scatter}
source("sources/discordant_analyses.R")
# Per-species with free limits — saves individual plots

invisible(plot_discordance_scatter(
  discordance_df, dep_raw, deg_raw, OUTPUT_PATH,
  save_per_species = TRUE
))

# Combined with shared limits
plots <- plot_discordance_scatter(
  discordance_df, dep_raw, deg_raw, OUTPUT_PATH,
  xlim = c(-11, 16), ylim = c(-5, 5)
)

combined <- wrap_plots(plots, nrow = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")

print(combined)
output_file <- file.path(OUTPUT_PATH, "discordance_scatter")
save_plot(combined, output_file, width = 16, height = 5)

message("Combined discordance scatter saved", output_file, MSG_FIG_FORMAT)
```
## Profile

```{r profile}
source("sources/discordant_analyses.R")

# Per-species
species_list <- unique(discordance_df$Species)
invisible(walk(species_list, function(species) {
  p <- plot_discordance_profiles(discordance_df, OUTPUT_PATH,
                                 species_filter = species
  )
  save_plot(
    plot = p,
    filepath = file.path(OUTPUT_PATH, species, "discordance_profiles"),
    width = 14, height = 5
  )
  message("Discordance profiles saved for: ", species, MSG_FIG_FORMAT)
}))

# Combined all species
combined <- plot_discordance_profiles(discordance_df, OUTPUT_PATH)
output_file <- file.path(OUTPUT_PATH, "discordance_profiles")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 20, height = 10
)
message("Combined discordance profiles saved: ", output_file, MSG_FIG_FORMAT)
print(combined)
```

## Enrichment

```{r enrichment-loading}
source("sources/discordant_analyses.R")

enrichment_results <- run_enrichment_all(
  discordance_df = discordance_df,
  pval_cutoff = 0.01,
  qval_cutoff = 0.2,
  output_path = OUTPUT_PATH
)
```

# Discordantly regulated genes are enriched in translation

```{r enrichment-mapping}
source("sources/discordant_analyses.R")
p <- plot_enrichment_dotplot(
  enrichment_results = enrichment_results,
  fdr_threshold = 0.05,
  top_n = 12, # limit to top 30 terms per cluster
  output_path = OUTPUT_PATH
)
print(p)
```

# mRNA-protein level stress-specific stimulon genes under stress[

```{r stimulon-data-generation}
# Code to generate stimulon specific data
```
## Stress specific stimulon genes and all genes mRNA-protein level correlation

```{r stimulon-correlation}

```
## ORA of osmotic stress-specific mRNA–protein associations across three species.
```{r ora-analysis}
```

# Supplementary figures
```{r supp-config}
SUPP_OUTPUT_PATH <- file.path(OUTPUT_PATH, "Supplementary")
dir.create(SUPP_OUTPUT_PATH)
for (species in SPECIES) dir.create(file.path(SUPP_OUTPUT_PATH, species))
```

## Protein and peptide lengths

```{r prot-length}
source("sources/protein_quantification.R")

fasta_lengths_df <- bind_rows(lapply(fasta_files, function(fp) {
  sp <- tools::file_path_sans_ext(
    stringr::str_extract(basename(fp), "^[A-Za-z]+_[A-Za-z]+"))
  fa <- Biostrings::readAAStringSet(fp)
  tibble::tibble(Species = sp, protein_length = Biostrings::width(fa))
}))
p1 <- plot_protein_length(fasta_lengths_df)

output_file <- file.path(SUPP_OUTPUT_PATH, "protein_length")
save_plot(
  plot = p1,
  filepath = output_file,
  width = 8, height = 8
)
message("  Protein lenght plot saved: ", output_file, MSG_FIG_FORMAT)

p2 <- plot_peptide_length(
  all_raw_digests,
  min_peptide_len = MIN_PEPTIDE_LEN,
  max_peptide_len = MAX_PEPTIDE_LEN
)
output_file <- file.path(SUPP_OUTPUT_PATH, "peptide_length")
save_plot(
  plot = p2,
  filepath = output_file,
  width = 8, height = 8
)
message(" Peptide plot saved: ", output_file, MSG_FIG_FORMAT)
fig <- p1 / p2 +
  patchwork::plot_annotation(
    theme = theme_publication() +
      theme(
        plot.title = element_text(size = 13, face = "bold"),
        plot.subtitle = element_text(size = 10, colour = "#5f5e5a",
                                     lineheight = 1.4)
      )
  ) &
  theme(plot.background = element_rect(fill = "white", colour = NA))

output_file <- file.path(SUPP_OUTPUT_PATH, "protein_digestion_analyses")
save_plot(
  plot = fig,
  filepath = output_file,
  width = 16, height = 16
)
message(" Protein digestion plot saved: ", output_file, MSG_FIG_FORMAT)
print(fig)
```

## Pairwise comparison of log2-transformed protein expression value (iBAQ) for each of the three replicates per condition.

Pearson correlation coefficient (R) and student-T test

```{r pairwise-comparison}
source("sources/protein_quantification.R")
# Guard: warn if log2_iBAQ has unexpected -Inf (zero iBAQ rows)
n_inf <- sum(is.infinite(protQ_data$iBAQ_log2))
if (n_inf > 0)
  message(sprintf(
    "Note: %d rows have -Inf log2_iBAQ (zero iBAQ). ",
    n_inf
  ), "These are excluded per scatter panel.")

species_list <- sort(unique(protQ_data$Species))
message(sprintf("Found %d species: %s",
                length(species_list),
                paste(species_list, collapse = ", ")))

walk(species_list, function(sp) {
  message("Process pairwise comparison for ", sp)
  sp_df <- protQ_data %>% filter(Species == sp)
  p <- plot_replicate_correlations(
    df = sp_df,
    species_name = sp
  )
  print(p)
  output_file <- file.path(SUPP_OUTPUT_PATH, sp, "replicate_correlation")
  save_plot(
    plot = p,
    filepath = output_file,
    width = 16, height = 16
  )
  message(" Pairwise comparison saved: ", output_file, MSG_FIG_FORMAT)

})
```
### Salmonella enterica

### Staphylococcus aureus

### Yersinia pseudotuberculosis

## Normalization
```{r normalization-comparison}
source("sources/normalization.R")
# Method comparison — generates NormalyzerDE reports and PCA plots

run_normalization_comparison(
  protQ_data,
  protQ = PROTEIN_QUANTIFICATION_METHOD,
  output_path = SUPP_OUTPUT_PATH
)
plot_pca_comparison_grid(
  protQ_data = protQ_data,
  norm_dir = file.path(SUPP_OUTPUT_PATH),
  output_path = SUPP_OUTPUT_PATH
)
```

## Essential DE
```{r de-essential-correlation}
source("sources/differential_expression.R")

# In read-DEG chunk, add after deg_raw is built
deg_raw_essential <- deg_raw %>%
  semi_join(essential_df, by = c("Species", "Protein_id"))

deg_data_essential <- deg_raw_essential %>%
  group_by(Species, Treatment) %>%
  summarise(
    n_de = sum(fdr < DEG_PVAL_THRESHOLD & abs(log2FC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = n(),
    pct_de = 100 * n_de / n_total,
    de_ids = list(Protein_id[fdr < DEG_PVAL_THRESHOLD &
                               abs(log2FC) > DE_FC_THRESHOLD &
                               !is.na(fdr)]),
    .groups = "drop"
  )

# In compute-DEP chunk, add Protein_id filter
dep_data_essential <- map_dfr(species_list, function(species) {
  sp_data <- norm_data %>%
    filter(
      Species == species,
      Treatment %in% c("Ctrl", STRESS_TREATMENTS_PROT),
      Protein_id %in% common_ids,
      Protein_id %in% essential_df$Protein_id[essential_df$Species == species],
      is.finite(value)
    )

  prot_expr <- build_expression_matrix(sp_data)

  run_limma_dep(
    mat = prot_expr$mat,
    metadata = prot_expr$metadata,
    species = species,
    pval_threshold = DEP_PVAL_THRESHOLD,
    fc_threshold = DE_FC_THRESHOLD
  )
})

plots <- plot_de_barplot_all(deg_data_essential, dep_data_essential, SUPP_OUTPUT_PATH, prefix = "Essential")
combined <- wrap_plots(plots, nrow = 1) +
  plot_layout(guides = "collect", axes = "collect_y") &
  theme(legend.position = "top")

print(combined)

output_file <- file.path(SUPP_OUTPUT_PATH, "Essential_DEG_DEP_barplot")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 16, height = 5
)
message("  Combined plot saved: ", output_file, MSG_FIG_FORMAT)

plots <- plot_de_vs_correlation_all(
  deg_data_essential,
  dep_data_essential,
  corr_essential_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD,
  SUPP_OUTPUT_PATH,
  prefix = "essential"
)
combined <- wrap_plots(plots, nrow = 2) +
  plot_layout(guides = "collect", axes = "collect_x") &
  theme(legend.position = "right")

print(combined)
output_file <- file.path(SUPP_OUTPUT_PATH, "Essential_DE_vs_correlation")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 16, height = 10
)
```

```{r discordant-scatter-per-treatment}
source("sources/discordant_analyses.R")
# Per-species with free limits — saves individual plots

for (treat in STRESS_TREATMENTS) {
  filtered_discordance_df <- discordance_df %>%
    filter(Treatment == treat)
  filtered_dep_raw <- dep_raw %>%
    filter(Treatment == treat)
  filtered_deg_raw <- deg_raw %>%
    filter(Treatment == treat)

  invisible(
    plot_discordance_scatter(
      filtered_discordance_df,
      filtered_dep_raw,
      filtered_deg_raw,
      SUPP_OUTPUT_PATH,
      prefix = treat,
      save_per_species = TRUE
    )
  )

  # Combined with shared limits
  plots <- plot_discordance_scatter(
    filtered_discordance_df,
    filtered_dep_raw,
    filtered_deg_raw,
    SUPP_OUTPUT_PATH,
    xlim = c(-11, 16), ylim = c(-5, 5)
  )

  combined <- wrap_plots(plots, nrow = 1) +
    plot_layout(guides = "collect") &
    theme(legend.position = "top")

  output_file <- file.path(
    SUPP_OUTPUT_PATH, paste(treat, "discordance_scatter", sep = "_")
  )
  save_plot(combined, output_file, width = 16, height = 5)

  message("Combined discordance scatter saved", output_file, MSG_FIG_FORMAT)
}
# Collect one combined plot per treatment
all_treatment_plots <- map(STRESS_TREATMENTS_PROT, function(treat) {
  filtered_discordance_df <- discordance_df %>% filter(Treatment == treat)
  filtered_dep_raw <- dep_raw %>% filter(Treatment == treat)
  filtered_deg_raw <- deg_raw %>% filter(Treatment == treat)

  plots <- plot_discordance_scatter(
    filtered_discordance_df,
    filtered_dep_raw,
    filtered_deg_raw,
    SUPP_OUTPUT_PATH,
    xlim = c(-11, 16), ylim = c(-5, 5)
  )

  # One combined row per treatment with title
  wrap_plots(plots, nrow = 1) +
    plot_layout(guides = "collect") +
    plot_annotation(title = treat) &
    theme(
      legend.position = "none", # hide per-treatment legends
      plot.title = element_text(face = "bold", size = 10)
    )
})

# Extract legend from one plot to place in the empty cell
legend_plot <- plot_discordance_scatter(
  discordance_df %>% filter(Treatment == STRESS_TREATMENTS_PROT[1]),
  dep_raw %>% filter(Treatment == STRESS_TREATMENTS_PROT[1]),
  deg_raw %>% filter(Treatment == STRESS_TREATMENTS_PROT[1]),
  SUPP_OUTPUT_PATH,
  xlim = c(-11, 16), ylim = c(-5, 5)
)[[1]] +
  theme(legend.position = "right") +
  guides(
    starshape = guide_legend(override.aes = list(fill = "grey40", colour = "grey40", size = 3, alpha = 1)),
    fill = guide_legend(override.aes = list(starshape = 15, colour = "white", size = 3, alpha = 1))
  )

# Build 4x3 grid: 10 treatments + 1 legend + 1 empty
grid_plots <- c(
  all_treatment_plots,
  list(cowplot::get_legend(legend_plot) %>% wrap_elements()),
  list(plot_spacer())
)

# One row per species across all treatments
walk(sort(unique(discordance_df$Species)), function(species) {
  species_plots <- map(STRESS_TREATMENTS_PROT, function(treat) {
    filtered_discordance_df <- discordance_df %>% filter(Treatment == treat, Species == species)
    filtered_dep_raw <- dep_raw %>% filter(Treatment == treat, Species == species)
    filtered_deg_raw <- deg_raw %>% filter(Treatment == treat, Species == species)

    plots <- plot_discordance_scatter(
      filtered_discordance_df,
      filtered_dep_raw,
      filtered_deg_raw,
      SUPP_OUTPUT_PATH,
      xlim = c(-11, 16), ylim = c(-5, 5)
    )

    plots[[1]] +
      theme(legend.position = "none") +
      labs(title = treat) +
      theme(plot.title = element_text(face = "bold", size = 9, hjust = 0.5))
  })

  # 4x3 grid with legend in last cell
  legend_p <- cowplot::get_legend(
    plots[[1]] +
      theme(legend.position = "right") +
      guides(
        starshape = guide_legend(override.aes = list(fill = "grey40", colour = "grey40", size = 3, alpha = 1)),
        fill = guide_legend(override.aes = list(starshape = 15, colour = "white", size = 3, alpha = 1))
      )
  )

  combined_species <- wrap_plots(
    c(species_plots, list(wrap_elements(legend_p), plot_spacer())),
    ncol = 3
  ) +
    plot_annotation(
      title = format_species_title(species),
      theme = theme(plot.title = element_text(hjust = 0.5, face = "italic", size = 12))
    )

  print(combined_species)
  output_file <- file.path(
    SUPP_OUTPUT_PATH, species,
    paste0("all_treatments_discordance_scatter")
  )
  save_plot(combined_species, output_file, width = 18, height = 24)
  message("All-treatment grid saved for: ", species, MSG_FIG_FORMAT)

  print(combined_species)
})
```

```{r full-enrichment-mapping, fig.height = 24}
source("sources/discordant_analyses.R")
p <- plot_enrichment_dotplot(
  enrichment_results = enrichment_results,
  fdr_threshold = 0.05,
  top_n = NULL, # limit to top 30 terms per cluster
  output_path = SUPP_OUTPUT_PATH
)
print(p)
```
# Annexes

## Generate DF for MOBILE

```{r generate-mobile-data}
raw_file <- file.path(OUTPUT_PATH, paste0(PROTEIN_QUANTIFICATION_METHOD, "_data.tsv"))
raw_data = read.table(
  raw_file,
  sep = "\t",
  header = TRUE
)
full_df <- raw_data %>%
  select(Protein_id, Species, Replicate, Treatment, all_of(PROTEIN_QUANTIFICATION_METHOD)) %>%
  mutate(sample = paste0(Replicate, "_", Treatment)) %>%
  select(-Replicate, -Treatment) %>%
  pivot_wider(names_from = sample, values_from = all_of(PROTEIN_QUANTIFICATION_METHOD)) %>%
  group_by(Protein_id) %>%
  mutate(ID = Protein_id) %>%
  select(-c(Protein_id)) %>%
  relocate(ID)

mean_df <- raw_data %>%
  select(Protein_id, Species, Replicate, Treatment, all_of(PROTEIN_QUANTIFICATION_METHOD)) %>%
  select(-Replicate) %>%
  pivot_wider(names_from = Treatment, values_from = all_of(PROTEIN_QUANTIFICATION_METHOD), values_fn = mean) %>%
  mutate(ID = Protein_id) %>%
  select(-c(Protein_id)) %>%
  relocate(ID)

full_df_list <- full_df %>%
  group_by(Species) %>%
  group_split(.keep = FALSE) %>%
  setNames(
    full_df %>%
      group_by(Species) %>%
      group_keys() %>%
      pull(Species)
  )

mean_df_list <- mean_df %>%
  group_by(Species) %>%
  group_split(.keep = FALSE) %>%
  setNames(
    mean_df %>%
      group_by(Species) %>%
      group_keys() %>%
      pull(Species)
  )

sp2abbv <- list(
  "Salmonella_enterica" = "SE",
  "Staphylococcus_aureus" = "SA",
  "Yersinia_pseudotuberculosis" = "YP"
)

iwalk(full_df_list, function(df, species) {
  output_file <- file.path("Analyses_iBAQ", paste0("iBAQ_data_", sp2abbv[[species]], ".tsv"))
  write.table(
    df,
    file = output_file,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
})

iwalk(mean_df_list, function(df, species) {
  output_file <- file.path("Analyses_iBAQ", paste0("iBAQ_data_mean_", sp2abbv[[species]], ".tsv"))
  write.table(
    df,
    file = output_file,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
})
```