---
title: "Bacterial Stress Responses Lower mRNA-Protein Level Correlations"
author: "Sena Gizem Suer, Yi Yang Lim, Rabia Şen, Cemal Erdem, André Mateus and Kemal Avican"
date: "2026-02-18"
output: 
    rmdformats::downcute:
        code_folding: hide
        self_contained: true
        thumbnails: true
        lightbox: true
pkgdown:
  as_is: true

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
    message = TRUE,
    warning = FALSE,
    error = FALSE
)
```

```{r load-environment}
# Load library
library(fs)
library(readxl)
library(tidyverse)
library(patchwork)
library(limma)

# Load sources functions
source("sources/utils.R")
```

```{r configuration, results='hide', error=TRUE}
#' Species Abbreviation Mapping
#' Maps full species names to abbreviations for data parsing
SPECIES <- c(
    "Salmonella_enterica",
    "Staphylococcus_aureus",
    "Yersinia_pseudotuberculosis"
)

SPECIES_ABBV_MAP <- c(
    "Salmonella_enterica"         = "SALMT",
    "Staphylococcus_aureus"       = "MSSA",
    "Yersinia_pseudotuberculosis" = "YPSTB"
)

#' Treatment List
#' Control and stress treatment conditions
treatment_list <- c(
    "Ctrl", "As", "Bs", "Mig", "Li", "Nd", "Ns", "Oss", "Oxs", "Sp", "Tm"
)

treatment_to_exposure_time <- c(
    "Ctrl" = 10,
    "As"   = 10,
    "Bs"   = 10,
    "Hyp"  = 240,
    "Li"   = 10,
    "Nd"   = 30,
    "Ns"   = 10,
    "Oss"  = 10,
    "Oxs"  = 10,
    "Sp"   = 241,
    "Tm"   = 20
)

#' Exposure Time Label Mapping
#' Maps numeric exposure times (minutes) to display labels
EXPOSURE_TIME_LABELS <- c(
    "10"  = "10",
    "20"  = "20 min",
    "30"  = "30 min",
    "240" = "4 hours",
    "241" = "\u22654 hours"
)

#' Measurement Display Labels
MEASUREMENT_LABELS <- c(
    "corr_intensity" = "Intensity",
    "corr_RI"        = "RI",
    "corr_iBAQ"      = "iBAQ",
    "corr_iBAQ_mc"   = "iBAQ MC"
)

#' Treatment Shape Mapping (ggstar)
TREATMENT_SHAPES <- c(
    "As"   = 1,
    "Bs"   = 14,
    "Ctrl" = 15,
    "Hyp"  = 11,
    "Li"   = 7,
    "Nd"   = 8,
    "Ns"   = 4,
    "Oss"  = 22,
    "Oxs"  = 21,
    "Sp"   = 23,
    "Tm"   = 2
)

# --- Directories ---------------------------------------------------------------
dirs <- list(
    DATA            = path("DATA"),
    RNA_DATA        = path("DATA", "Read_counts"),
    PROTEIN_DATA    = path("DATA", "Proteome"),
    PROTEIN_FASTA   = path("DATA", "Fasta_files"),
    ESSENTIAL_GENES = path("DATA", "Essential genes")
)

walk(dirs, check_dir, return_error = TRUE)

OUTPUT_PATH <- path("Analyses2")
dir_create(OUTPUT_PATH)
for (species in SPECIES) dir_create(path(OUTPUT_PATH, species))

# --- Required files ------------------------------------------------------------
files <- list(
    SAMPLE_MAPPING = path(dirs$PROTEIN_DATA, "sample_mapping.xlsx")
)

walk(files, ~ if (!file_exists(.x)) stop(.x, " not found"))

# --- Expose to environment (equivalent to your original variables) -------------
list2env(c(dirs, files), envir = .GlobalEnv)
```

# Data loading and processing
```{r load-tpm, results='hide'}
source("sources/read_data.R")
# Load RNA tpm data
rna_files <- list.files(RNA_DATA, pattern = "\\.xlsx$", full.names = TRUE)
tpm_data <- list()
lapply(rna_files, function(rna_file) {
    species <- get_species_tpm(rna_file)
    abbv <- SPECIES_ABBV_MAP[species]
    tpm_raw <- read_tpm(rna_file, abbv) %>%
        mutate(
            Species = species,
            Treatment = ifelse(Treatment == "Mig", "Hyp", Treatment),
            Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE)
        ) %>%
        filter(!is.na(Protein_id), !is.na(TPM)) %>%
        log2_transform(value_col = "TPM", pseudocount = TRUE) %>%
        filter(is.finite(mean_log2)) %>%
        rename(mean_log2_TPM = mean_log2)

    if (!is.null(tpm_raw)) {
        tpm_data[[species]] <<- tpm_raw
    }
})
```

```{r load-prot, results='hide'}
# source("sources/read_data.R")
# Load proteomic data
sample_mapping <- read_excel(SAMPLE_MAPPING, col_names = TRUE) %>%
    mutate(Replicate = paste0("R", Replicate))
protein_files <- list.files(
    PROTEIN_DATA,
    pattern = "\\.tsv$",
    recursive = TRUE,
    full.names = TRUE
)

filter_log <- list()
filtered_proteins <- list()

lapply(protein_files, function(protein_file) {
    species <- sub(".*/(\\w+)_R\\d+\\.tsv$", "\\1", protein_file)
    replicate <- sub(".*/\\w+_(R\\d+)\\.tsv$", "\\1", protein_file)

    tmt_raw <- read_tmt(protein_file)
    filtered_raw <- filter_tmt(tmt_raw, species, replicate)

    filtered_proteins[[species]][[replicate]] <<- filtered_raw$filtered_tmt
    filter_log[[species]][[replicate]] <<- filtered_raw$filter_log
})

tmt_data <- lapply(names(filtered_proteins), function(species) {
    map_dfr(names(filtered_proteins[[species]]), function(rep) {
        filtered_proteins[[species]][[rep]] %>%
            mutate(Species = species, Replicate = rep)
    })
}) %>% set_names(names(filtered_proteins))

# TODO put this in a function
# Build and write filter log
filter_log_df <- map_dfr(filter_log, ~ bind_rows(.x)) %>%
    mutate(across(-c("species", "replicate"), as.integer))

message("TMT filtering summary:\n")

write.table(
    filter_log_df,
    path(OUTPUT_PATH, "filter_log.tsv"),
    sep       = "\t",
    row.names = FALSE,
    quote     = FALSE
)
```

# Trancriptomic and Proteomic analysis of 3 human pathogens under 10 revelevant stress condition

```{r load-annotation, results='hide'}
annotation_files <- list.files(
    PROTEIN_DATA,
    pattern     = "^annotation\\.xlsx$", # ^ and $ anchor to exact filename
    recursive   = TRUE,
    full.names  = TRUE
) # %>%
# .[!grepl("^~\\$", basename(.))]        # exclude Excel temp files (~$annotation.xlsx)
annotation_data <- list()
lapply(annotation_files, function(annotation_file) {
    species <- path_file(path_dir(annotation_file))
    annotation_raw <- read_excel(
        annotation_file,
        col_names = TRUE, na = "N/A"
    ) %>%
        select(locus_tag, old_locus_tag) %>%
        mutate(
            Protein_id = if_else(
                is.na(locus_tag),
                old_locus_tag,
                locus_tag
            )
        ) %>%
        mutate(
            Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE)
        )
    annotation_data[[species]] <<- annotation_raw
})
```

## Percentages of CDSs detected by transcriptomics and proteomics for the three species 
```{r cds-distribution}
source("sources/cds_distribution.R")

detected_cds_df <- get_detected_cds(annotation_data, tpm_data, tmt_data)
detected_cds_file <- path(OUTPUT_PATH, "Detected_CDS.tsv")
write.table(detected_cds_df, detected_cds_file, sep = "\t")
plot_barplot_detected_cds(
    detected_cds_df,
    position = "overlay", path(OUTPUT_PATH, "CDS_detected.png")
)
```

## Distribution of transcript abundance across all conditions in the three species

- All condition: light blue,
- Fraction of proteins detected: yellow,
- Transcripts with no corresponding protein detected: dark blue.

```{r plot-transcript-abundance}
# Shared x_max across all species for visual consistency
source("sources/cds_distribution.R")
x_max <- map_dfr(tpm_data, ~.x) %>%
    summarise(x_max = ceiling(max(mean_log2_TPM, na.rm = TRUE) / 2.5) * 2.5) %>%
    pull(x_max)
    
plots <- imap(tpm_data, function(tpm_sp, species) {
    message("Plotting TPM detection for: ", species)

    classified <- classify_tpm_groups(tpm_sp, tmt_data[[species]])

    p <- plot_tpm_detection_species(classified, species, x_max = x_max)

    output_sp <- path(OUTPUT_PATH, species, "TPM_detection_distribution")
    save_plot(
        plot = p,
        filepath = path(OUTPUT_PATH, species, "TPM_detection_distribution"),
        width = 5,
        height = 4
    )

    message("Combined figure saved: ", output_sp, ".pdf/.png")

    p
})

# Combined figure — all species side by side
combined <- wrap_plots(plots, nrow = 1) +
    plot_layout(guides = "collect") &
    theme(legend.position = "right")

print(combined)

save_plot(
    plot     = combined,
    filepath = file.path(OUTPUT_PATH, "AllSpecies_TPM_detection_distribution"),
    width    = 5 * length(plots),
    height   = 4
)

message("Combined figure saved: AllSpecies_TPM_detection_distribution.pdf/.png")
```

# mRNA and protein level correlation under stress


