---
title: "Bacterial Stress Responses Lower mRNA-Protein Level Correlations"
author: "Sena Gizem Suer, Yi Yang Lim, Rabia Şen, Cemal Erdem, André Mateus and Kemal Avican"
date: "2026-02-18"
output:
  rmdformats::downcute:
    code_folding: hide
    self_contained: true
    thumbnails: false
    lightbox: true
pkgdown:
  as_is: true
---

# How to use

## Requirements

- **R ≥ 4.4** — download from <https://cran.r-project.org>
- **Rtools** (Windows only) — must match your R version exactly.
  Download from <https://cran.r-project.org/bin/windows/Rtools/>

## 1 — Clone the repository

```bash
git clone https://github.com/<org>/<repo>.git
cd <repo>
```

## 2 — Restore the R environment with renv

This project uses [renv](https://rstudio.github.io/renv/) to pin all package
versions. Open R (or RStudio) in the project root and run:

```r
install.packages("renv")
renv::restore()
```

**Common issues**

| Problem                                          | Solution                                                                                                                      |
|--------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------|
| `BiocParallel` fails to compile                  | Update Rtools to match your R version, then re-run `renv::restore()`                                                          |
| A single package cannot be installed             | Exclude it: `renv::restore(exclude = "pkgName")`, then install manually with `install.packages()` or `BiocManager::install()` |
| `renv::restore()` asks to activate — say **yes** | This links the project library so packages are isolated                                                                       |

## 3 — Download data from Zenodo

All raw data files are archived on Zenodo at
**<https://doi.org/10.5281/zenodo.XXXXXXX>**.

### Option A — R (recommended, reproducible)

```r
# install.packages("zen4R")   # run once if not already installed
library(zen4R)

zenodo <- ZenodoManager$new()
record <- zenodo$getRecordByDOI("10.5281/zenodo.XXXXXXX")
zenodo$downloadFiles(record, path = "DATA")
```

### Option B — Command line

```bash
# requires curl ≥ 7.x
DOI_URL="https://zenodo.org/api/records/XXXXXXX/files-archive"
curl -L "$DOI_URL" -o zenodo_data.zip
unzip zenodo_data.zip -d DATA/
```

### Option C — Browser

1. Open <https://doi.org/10.5281/zenodo.XXXXXXX>
2. Click **Download all** (or download individual files)
3. Unzip into the `DATA/` folder at the project root

### Expected DATA folder structure

```
DATA/
├── Read_counts/          # RNA-seq TPM Excel files  (*.xlsx)
├── Proteome/             # TMT result TSV files, annotation, sample mapping
│   ├── sample_mapping.xlsx
│   ├── <Species>/
│   │   ├── annotation.xlsx
│   │   └── <Species>_R<n>.tsv
├── Fasta_files/          # Proteome FASTA files     (*.fa)
├── Essential genes/      # Essential gene lists     (*.xlsx)
├── Enrichment/           # GO term files            (*.xlsx / *.csv)
├── Stimulons/            # Stimulon gene lists      (*.xlsx)
└── ORA/                  # MOBILE framework output  (*.csv / *.txt)
```

## 4 — Render the notebook

```r
rmarkdown::render("main.Rmd")
```

Or use the **Knit** button in RStudio.

---

# Configuration

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message   = FALSE,
  warning   = FALSE,
  error     = FALSE,
  fig.height = 6,
  fig.width  = 12
)
```

## Load libraries

```{r load-environment, results='hide'}
library(fs)
library(purrr)
library(patchwork)
library(cowplot)
library(stringr)
library(tidyverse)
library(readxl)
library(sysfonts)   # font_add() — must be loaded before showtext
library(showtext)
library(limma)
library(openxlsx)
library(clusterProfiler)

source("sources/utils.R")
```

## Paths and global variables

```{r configuration, results='hide', error=TRUE}
# ── Species ────────────────────────────────────────────────────────────────────
SPECIES <- c(
  "Salmonella_enterica",
  "Staphylococcus_aureus",
  "Yersinia_pseudotuberculosis"
)

SPECIES_ABBV_MAP <- c(
  "Salmonella_enterica"       = "SALMT",
  "Staphylococcus_aureus"     = "MSSA",
  "Yersinia_pseudotuberculosis" = "YPSTB"
)

PLOT_SPECIES_NAMES <- c(
  "Salmonella_enterica"       = "S. enterica Typhimurium",
  "Staphylococcus_aureus"     = "S. aureus",
  "Yersinia_pseudotuberculosis" = "Y. pseudotuberculosis"
)

# ── Treatments ─────────────────────────────────────────────────────────────────
STRESS_TREATMENTS      <- c("As", "Bs", "Li", "Nd", "Ns", "Oss", "Oxs", "Sp", "Tm")
STRESS_TREATMENTS_RNA  <- c(STRESS_TREATMENTS, "Mig")
STRESS_TREATMENTS_PROT <- c(STRESS_TREATMENTS, "Hyp")
TREATMENTS_LIST_RNA    <- c("Ctrl", STRESS_TREATMENTS_RNA)

# ── Display labels ─────────────────────────────────────────────────────────────

# Maps numeric exposure times (minutes) to display labels
EXPOSURE_TIME_LABELS <- c(
  "10"  = "10 min",
  "20"  = "20 min",
  "30"  = "30 min",
  "240" = "4 hours",
  "241" = "\u22654 hours"   # ≥4 hours
)

MEASUREMENT_LABELS <- c(
  "corr_intensity" = "Intensity",
  "corr_RI"        = "RI",
  "corr_iBAQ"      = "iBAQ",
  "corr_iBAQ_mc"   = "iBAQ MC"
)

# ggstar shape codes per treatment
TREATMENT_SHAPES <- c(
  "As"  = 1,  "Bs"  = 14, "Ctrl" = 15, "Hyp" = 11,
  "Li"  = 7,  "Nd"  = 8,  "Ns"   = 4,  "Oss" = 22,
  "Oxs" = 21, "Sp"  = 23, "Tm"   = 2
)

TREATMENT_COLOR <- c(
  "As" = "#F8766D", "Bs" = "#D89000", "Hyp" = "#A3A500",
  "Li" = "#39B600", "Nd" = "#00BF7D", "Ns" = "#00BFC4",
  "Oss" = "#00B0F6", "Oxs" = "#9590FF", "Sp" = "#E76BF3",
  "Tm" = "#FF62BC", "Ctrl" = "black"
)


# ── Protein quantification method ──────────────────────────────────────────────
# Uncomment the desired method (only one should be active):
# PROTEIN_QUANTIFICATION_METHOD <- "Intensity"  # Direct TMT channel intensity
# PROTEIN_QUANTIFICATION_METHOD <- "RI"         # Intensity / razor peptides
PROTEIN_QUANTIFICATION_METHOD <- "iBAQ"         # Intensity / expected tryptic peptides
# PROTEIN_QUANTIFICATION_METHOD <- "iBAQ_MC"    # iBAQ with missed-cleavage filter

# ── Directories ────────────────────────────────────────────────────────────────
dirs <- list(
  DATA           = file.path("DATA"),
  RNA_DATA       = file.path("DATA", "Read_counts"),
  PROTEIN_DATA   = file.path("DATA", "Proteome"),
  PROTEIN_FASTA  = file.path("DATA", "Fasta_files"),
  ESSENTIAL_GENES = file.path("DATA", "Essential genes"),
  ENRICHMENT     = file.path("DATA", "Enrichment"),
  STIMULON       = file.path("DATA", "Stimulons"),
  ORA            = file.path("DATA", "ORA")
)

purrr::walk(dirs, check_dir, return_error = TRUE)

# ── Output paths ───────────────────────────────────────────────────────────────
OUTPUT_PATH <- file.path("Analyses_iBAQ")
FIGURE_FORMAT <- c("png", "svg", "pdf")
MSG_FIG_FORMAT <- paste0(".", paste(FIGURE_FORMAT, collapse = "/."))

dir.create(OUTPUT_PATH, showWarnings = FALSE)
purrr::walk(SPECIES, ~ dir.create(file.path(OUTPUT_PATH, .x), showWarnings = FALSE))

# ── Required files ─────────────────────────────────────────────────────────────
files <- list(
  SAMPLE_MAPPING = file.path(dirs$PROTEIN_DATA, "sample_mapping.xlsx")
)
purrr::walk(files, ~ if (!fs::file_exists(.x)) stop(.x, " not found"))

# Expose directories and files to the global environment for convenience
list2env(c(dirs, files), envir = .GlobalEnv)

# ── Fonts ──────────────────────────────────────────────────────────────────────
# font_add() lives in sysfonts (a dependency of showtext), not in showtext itself,
# so it cannot be called with showtext::font_add().
# We also guard the registration so the notebook renders on systems where
# arial.ttf is absent (e.g. Linux CI), falling back to the default sans-serif.
if (nzchar(Sys.which("arial.ttf")) || file.exists("arial.ttf")) {
  sysfonts::font_add(family = "Arial", regular = "arial.ttf")
} else {
  message("arial.ttf not found — using system sans-serif as 'Arial' fallback.")
  sysfonts::font_add(family = "Arial", regular = sysfonts::font_files()$path[1])
}

if (.Platform$OS.type == "windows") {
  grDevices::windowsFonts(
    mono  = grDevices::windowsFont("Courier New"),
    Arial = grDevices::windowsFont("Arial")
  )
}
showtext::showtext_auto()
```

---

# Results

## Data loading and processing

```{r load-tpm, results='hide'}
source("sources/read_data.R")

rna_files <- list.files(RNA_DATA, pattern = "\\.xlsx$", full.names = TRUE)

tpm_data <- purrr::map_dfr(rna_files, function(rna_file) {
  species <- get_species_tpm(rna_file)
  abbv    <- SPECIES_ABBV_MAP[species]
  read_tpm(rna_file, abbv) |>
    dplyr::mutate(
      # Harmonise treatment label: Mig → Hyp (migration = hyperosmotic shock)
      Treatment  = ifelse(Treatment == "Mig", "Hyp", Treatment),
      # Fix capitalisation: "pi..." → "PI..."
      Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE),
      Species    = species
    ) |>
    dplyr::filter(!is.na(Protein_id), !is.na(TPM))
})

# Remove genes with zero TPM across all conditions and replicates
zero_tpm_ids <- tpm_data |>
  dplyr::group_by(Protein_id) |>
  dplyr::summarise(TPM_sum = sum(TPM), .groups = "drop") |>
  dplyr::filter(TPM_sum == 0) |>
  dplyr::pull(Protein_id)

message("Removing ", length(zero_tpm_ids), " genes with TPM = 0 across all conditions")
tpm_data <- dplyr::filter(tpm_data, !Protein_id %in% zero_tpm_ids)
```

```{r load-prot, results='hide'}
source("sources/read_data.R")

sample_mapping <- readxl::read_excel(SAMPLE_MAPPING, col_names = TRUE) |>
  dplyr::mutate(Replicate = paste0("R", Replicate))

protein_files <- list.files(
  PROTEIN_DATA,
  pattern   = "\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)

# Read, filter, and annotate all TMT files; collect per-file filter logs
filter_log <- list()

filtered_data <- purrr::map_dfr(protein_files, function(protein_file) {
  species   <- sub(".*/(\\w+)_R\\d+\\.tsv$",  "\\1", protein_file)
  replicate <- sub(".*/\\w+_(R\\d+)\\.tsv$",   "\\1", protein_file)

  tmt_raw     <- read_tmt(protein_file)
  filtered_raw <- filter_tmt(tmt_raw, species, replicate)

  filter_log[[species]][[replicate]] <<- filtered_raw$filter_log

  dplyr::mutate(filtered_raw$filtered_tmt, Species = species, Replicate = replicate)
}) |>
  dplyr::left_join(sample_mapping, by = c("Species", "Replicate"), relationship = "many-to-many")

# Compile and save the filter log
filter_log_df <- purrr::map_dfr(filter_log, ~ dplyr::bind_rows(.x)) |>
  dplyr::mutate(dplyr::across(-c("species", "replicate"), as.integer))

message("TMT filtering summary:")
utils::write.table(
  filter_log_df,
  file.path(OUTPUT_PATH, "filter_log.tsv"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

# Keep only proteins observed in every Replicate × Treatment combination
completeness_result <- filter_complete_proteins(filtered_data)
tmt_data            <- completeness_result$filtered_data
removed_proteins    <- completeness_result$removed_proteins

if (nrow(removed_proteins) > 0) {
  utils::write.table(
    dplyr::arrange(removed_proteins, Species, Protein_id),
    file.path(OUTPUT_PATH, "removed_proteins_incomplete.tsv"),
    sep = "\t", row.names = FALSE, quote = FALSE
  )
  message("Removed proteins written to: removed_proteins_incomplete.tsv")
}
```

# Transcriptomic and proteomic analysis of 3 human pathogens under 10 relevant stress conditions

```{r load-annotation, results='hide'}
annotation_files <- list.files(
  PROTEIN_DATA,
  pattern   = "^annotation\\.xlsx$",
  recursive = TRUE,
  full.names = TRUE
)

annotation_data <- purrr::map_dfr(annotation_files, function(annotation_file) {
  species <- fs::path_file(fs::path_dir(annotation_file))
  readxl::read_excel(annotation_file, col_names = TRUE, na = "N/A") |>
    dplyr::select(locus_tag, old_locus_tag) |>
    dplyr::mutate(
      Protein_id = dplyr::if_else(is.na(locus_tag), old_locus_tag, locus_tag),
      Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE),
      Species    = species
    )
})
```

## Percentages of CDSs detected by transcriptomics and proteomics

```{r cds-distribution}
source("sources/cds_distribution.R")

detected_cds_df <- get_detected_cds(annotation_data, tpm_data, tmt_data)
print(detected_cds_df)
utils::write.table(
  detected_cds_df,
  file.path(OUTPUT_PATH, "Detected_CDS.tsv"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

p <- plot_barplot_detected_cds(detected_cds_df, position = "overlay")
print(p)

output_file <- file.path(OUTPUT_PATH, "CDS_detected")
save_plot(p, output_file, width = 6, height = 9)
message("CDS detected plot saved: ", output_file, MSG_FIG_FORMAT)
```

## Distribution of transcript abundance across all conditions

*Colours: all mRNA (light blue) · proteins detected (yellow) · mRNA only — no protein detected (dark blue).*

```{r plot-transcript-abundance}
source("sources/cds_distribution.R")

tpm_log2_data <- tpm_data |>
  log2_transform(
    value_col  = "TPM",
    group_cols = c("Species", "Protein_id", "Treatment"),
    pseudocount = TRUE
  ) |>
  dplyr::filter(is.finite(mean_log2)) |>
  dplyr::rename(mean_log2_TPM = mean_log2)

# Shared x-axis maximum across species for visual consistency
x_max <- tpm_log2_data |>
  dplyr::summarise(
    x_max = ceiling(max(mean_log2_TPM, na.rm = TRUE) / 2.5) * 2.5
  ) |>
  dplyr::pull(x_max)

species_list <- sort(unique(tpm_log2_data$Species))

plots <- purrr::imap(species_list, function(species, i) {
  message("Plotting TPM detection for: ", species)
  classified <- classify_tpm_groups(tpm_log2_data, tmt_data, species)
  p          <- plot_tpm_detection_species(classified, species, x_max = x_max)

  output_sp <- file.path(OUTPUT_PATH, species, "TPM_detection_distribution")
  save_plot(p, output_sp, width = 12, height = 6)
  message("CDS distribution for ", species, " saved: ", output_sp, MSG_FIG_FORMAT)

  # Remove redundant y-axis titles from panels 2 and 3
  if (i > 1) p <- p + ggplot2::theme(axis.title.y = ggplot2::element_blank())
  p
})

combined <- patchwork::wrap_plots(plots, nrow = 1) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "top")

print(combined)

output_file <- file.path(OUTPUT_PATH, "TPM_detection_distribution")
save_plot(combined, output_file, width = 12, height = 6)
message("Combined TPM distribution figure saved: ", output_file, MSG_FIG_FORMAT)
```

---

# mRNA and protein level correlation under stress

## Protein quantification

```{r in-silico-digest, eval=PROTEIN_QUANTIFICATION_METHOD %in% c("iBAQ", "iBAQ_MC")}
# In silico trypsin digestion — only required for iBAQ methods
MIN_PEPTIDE_LEN    <- 6
MAX_PEPTIDE_LEN    <- 30
IBAQ_MISS_CLEAVAGE <- 2

source("sources/protein_quantification.R")

fasta_files <- list.files(PROTEIN_FASTA, pattern = "\\.fa$", full.names = TRUE)
results <- process_tryptic(
  fasta_paths      = fasta_files,
  output_path      = OUTPUT_PATH,
  min_peptide_len  = MIN_PEPTIDE_LEN,
  max_peptide_len  = MAX_PEPTIDE_LEN,
  missed_cleavages = ifelse(PROTEIN_QUANTIFICATION_METHOD == "iBAQ", 0, IBAQ_MISS_CLEAVAGE)
)
protein2tryptic <- results$observable_peptides
all_raw_digests <- results$all_raw_digests
```

```{r protein-quantification}
source("sources/protein_quantification.R")

message("Quantification method: ", PROTEIN_QUANTIFICATION_METHOD)

# Dispatch table: method name → quantification function
quant_fn <- switch(
  PROTEIN_QUANTIFICATION_METHOD,
  RI        = function(x) ri_quantification(x),
  Intensity = function(x) intensity_quantification(x),
  iBAQ      = function(x) ibaq_quantification(x, protein2tryptic, "n_peptides", OUTPUT_PATH),
  iBAQ_MC   = function(x) ibaq_quantification(x, protein2tryptic, "n_peptides", OUTPUT_PATH),
  stop("Unknown quantification method: ", PROTEIN_QUANTIFICATION_METHOD)
)

quant_data         <- quant_fn(tmt_data)
completeness_result <- filter_complete_proteins(quant_data, PROTEIN_QUANTIFICATION_METHOD)
protQ_data          <- completeness_result$filtered_data
removed_proteins    <- completeness_result$removed_proteins

if (nrow(removed_proteins) > 0) {
  removed_path <- file.path(
    OUTPUT_PATH,
    paste0(PROTEIN_QUANTIFICATION_METHOD, "_removed_proteins_incomplete.tsv")
  )
  utils::write.table(
    dplyr::arrange(removed_proteins, Species, Protein_id),
    removed_path, sep = "\t", row.names = FALSE, quote = FALSE
  )
  message(PROTEIN_QUANTIFICATION_METHOD, " removed proteins written to: ", removed_path)
}

protQ_path <- file.path(OUTPUT_PATH, paste0(PROTEIN_QUANTIFICATION_METHOD, "_data.tsv"))
utils::write.table(protQ_data, protQ_path, sep = "\t", row.names = FALSE, quote = FALSE)
message(PROTEIN_QUANTIFICATION_METHOD, " quantification data saved: ", protQ_path)
```

## mRNA-protein correlation across stress and control conditions

```{r rna-protein-correlation, fig.height=18}
source("sources/correlation.R")

protQ_meanlog2 <- paste0(PROTEIN_QUANTIFICATION_METHOD, "_meanlog2")
common_ids     <- extract_common_ids(tpm_log2_data, protQ_data)

tpm_log2_data <- dplyr::filter(tpm_log2_data, Protein_id %in% common_ids)
protQ_data    <- dplyr::filter(protQ_data,    Protein_id %in% common_ids)

corr_rna_prot <- calculate_correlation(tpm_log2_data, protQ_data, protQ_meanlog2)

p <- plot_correlation(corr_rna_prot, protQ_meanlog2, PROTEIN_QUANTIFICATION_METHOD, OUTPUT_PATH)
print(p)

output_file <- file.path(OUTPUT_PATH, paste0("TPM_", PROTEIN_QUANTIFICATION_METHOD, "_correlation_panels"))
save_plot(p, output_file, width = 12, height = 18, units = "in")
message("Combined correlation figure saved: ", output_file, MSG_FIG_FORMAT)
```

---

# Essential genes have higher mRNA-protein level correlations

```{r get-essential-data}
source("sources/essential_analyses.R")

# Helper: log2-transform and compute SD in one step, keeping both columns
log2_mean_sd_transform <- function(data, value_col,
                                   group_cols  = c("Species", "Protein_id"),
                                   pseudocount = FALSE) {
  mean_col <- paste0("mean_log2_", value_col)
  sd_col   <- paste0("sd_log2_",   value_col)

  mean_data <- log2_transform(data, value_col,
                              group_cols  = group_cols,
                              pseudocount = pseudocount) |>
    dplyr::filter(is.finite(mean_log2)) |>
    dplyr::rename(!!mean_col := mean_log2)

  sd_data <- log2_mean_sd(data, value_col, sd_col,
                          group_cols  = group_cols,
                          pseudocount = pseudocount)

  dplyr::left_join(mean_data, sd_data, by = group_cols)
}

essential_files <- list.files(ESSENTIAL_GENES, pattern = "\\.xlsx$", full.names = TRUE)

essential_df <- purrr::map_dfr(essential_files, function(file) {
  species <- stringr::str_extract(basename(file), "Essential_genes_([A-Za-z]+)", group = 1)
  tibble::tibble(
    Species    = SPECIES_ESSENTIAL_TO_TPM_MAP[species],
    Protein_id = read_essential_vec(file)
  )
})

tpm_glob_log2mean_data   <- log2_mean_sd_transform(
  dplyr::filter(tpm_data,   Protein_id %in% common_ids),
  "TPM", pseudocount = TRUE
)
protQ_glob_log2mean_data <- log2_mean_sd_transform(
  dplyr::filter(protQ_data, Protein_id %in% common_ids),
  PROTEIN_QUANTIFICATION_METHOD
)

# Summary: how many essential genes are detected in each omic layer
essential_counts_table <- essential_df |>
  dplyr::group_by(Species) |>
  dplyr::summarise(n_essential_total = dplyr::n_distinct(Protein_id), .groups = "drop") |>
  dplyr::left_join(
    dplyr::semi_join(essential_df, tpm_glob_log2mean_data,   by = c("Species", "Protein_id")) |>
      dplyr::group_by(Species) |>
      dplyr::summarise(n_essential_rna  = dplyr::n_distinct(Protein_id), .groups = "drop"),
    by = "Species"
  ) |>
  dplyr::left_join(
    dplyr::semi_join(essential_df, protQ_glob_log2mean_data, by = c("Species", "Protein_id")) |>
      dplyr::group_by(Species) |>
      dplyr::summarise(n_essential_prot = dplyr::n_distinct(Protein_id), .groups = "drop"),
    by = "Species"
  )

print(essential_counts_table)
```

## Mean mRNA expression — essential vs non-essential genes

```{r essential-distribution-rna}
source("sources/essential_analyses.R")

p <- plot_essential_distribution(
  all_df       = tpm_glob_log2mean_data,
  essential_df = essential_df,
  log2_col     = "mean_log2_TPM",
  x_lab        = "TPM",
  output_path  = OUTPUT_PATH
)
print(p)

output_file <- file.path(OUTPUT_PATH, "TPM_essential_distribution")
save_plot(p, output_file, width = 12, height = 6, units = "in")
message("TPM essential distribution figure saved: ", output_file, MSG_FIG_FORMAT)
```

## Mean protein expression — essential vs non-essential genes

```{r essential-distribution-prot}
source("sources/essential_analyses.R")

p <- plot_essential_distribution(
  all_df       = protQ_glob_log2mean_data,
  essential_df = essential_df,
  log2_col     = paste0("mean_log2_", PROTEIN_QUANTIFICATION_METHOD),
  x_lab        = PROTEIN_QUANTIFICATION_METHOD,
  output_path  = OUTPUT_PATH
)
print(p)

output_file <- file.path(OUTPUT_PATH, paste0(PROTEIN_QUANTIFICATION_METHOD, "_essential_distribution"))
save_plot(p, output_file, width = 12, height = 6, units = "in")
message(PROTEIN_QUANTIFICATION_METHOD, " essential distribution figure saved: ", output_file, MSG_FIG_FORMAT)
```

## Essential gene mRNA-protein correlation vs all-gene correlation

mRNA-protein correlations were measured with the Pearson coefficient (*R*).
P-values were calculated using a Wilcoxon rank-sum test.

```{r essential-correlation}
source("sources/essential_analyses.R")

essential_tpm_df   <- dplyr::semi_join(tpm_log2_data, essential_df, by = c("Species", "Protein_id")) |>
  dplyr::filter(is.finite(mean_log2_TPM))
essential_protQ_df <- dplyr::semi_join(protQ_data, essential_df,    by = c("Species", "Protein_id")) |>
  dplyr::filter(is.finite(.data[[protQ_meanlog2]]))

corr_essential_rna_prot <- calculate_correlation(
  essential_tpm_df, essential_protQ_df, protQ_meanlog2
)

plots <- plot_subset_vs_all_correlation(
  corr_rna_prot, corr_essential_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD, "essential", OUTPUT_PATH
)

combined <- patchwork::wrap_plots(plots, ncol = 3) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "top")

print(combined)

output_file <- file.path(OUTPUT_PATH, "essential_vs_all_correlation")
save_plot(combined, output_file, width = 12, height = 4)
message("Essential vs all correlation figure saved: ", output_file, MSG_FIG_FORMAT)
```

## SD of mRNA expression — essential vs non-essential genes

```{r essential-sd-rna}
source("sources/essential_analyses.R")

p <- plot_sd_distribution(
  all_df       = tpm_glob_log2mean_data,
  essential_df = essential_df,
  sd_col       = "sd_log2_TPM",
  x_lab        = "TPM",
  output_path  = OUTPUT_PATH
)
print(p)

output_file <- file.path(OUTPUT_PATH, "TPM_essential_sd_distribution")
save_plot(p, output_file, width = 12, height = 6, units = "in")
message("TPM SD distribution figure saved: ", output_file, MSG_FIG_FORMAT)
```

## SD of protein expression — essential vs non-essential genes

```{r essential-sd-prot}
source("sources/essential_analyses.R")

p <- plot_sd_distribution(
  all_df       = protQ_glob_log2mean_data,
  essential_df = essential_df,
  sd_col       = paste0("sd_log2_", PROTEIN_QUANTIFICATION_METHOD),
  x_lab        = PROTEIN_QUANTIFICATION_METHOD,
  output_path  = OUTPUT_PATH
)
print(p)

output_file <- file.path(OUTPUT_PATH, paste0(PROTEIN_QUANTIFICATION_METHOD, "_essential_sd_distribution"))
save_plot(p, output_file, width = 12, height = 6, units = "in")
message(PROTEIN_QUANTIFICATION_METHOD, " SD distribution figure saved: ", output_file, MSG_FIG_FORMAT)
```

---

# Stress responses differentially affect mRNA-protein correlation

## Correlation vs stress exposure time

```{r exposure-time}
source("sources/correlation.R")

# Maps each treatment to its exposure duration in minutes.
# Sp is coded as 241 (> 240) to represent "≥ 4 hours".
treatment_to_exposure_time <- c(
  "As"  = 10,  "Bs"  = 10,  "Hyp" = 240,
  "Li"  = 10,  "Nd"  = 30,  "Ns"  = 10,
  "Oss" = 10,  "Oxs" = 10,  "Sp"  = 241,
  "Tm"  = 20
)

exposure_corr_df <- corr_rna_prot |>
  dplyr::filter(Treatment != "Ctrl") |>
  dplyr::distinct(Species, Treatment, R) |>
  dplyr::mutate(
    exposure_time  = treatment_to_exposure_time[Treatment],
    exposure_label = factor(
      EXPOSURE_TIME_LABELS[as.character(exposure_time)],
      levels = EXPOSURE_TIME_LABELS
    )
  )

stats_df  <- compute_exposure_correlations(exposure_corr_df)
label_pos <- build_label_position(exposure_corr_df, stats_df)

p <- build_exposure_plot(df = exposure_corr_df, label_pos = label_pos)
print(p)

output_file <- file.path(OUTPUT_PATH, "correlation_vs_exposure_time")
save_plot(p, output_file, width = 12, height = 5, units = "in")
message("Correlation vs exposure time figure saved: ", output_file, MSG_FIG_FORMAT)
```

## Protein normalization

```{r normalization}
source("sources/normalization.R")

# VSN is used for the main analysis; see the Supplementary section for a
# full comparison of all NormalyzerDE methods.
NORMALIZATION_METHOD <- "VSN"

norm_data <- run_normalization(
  protQ_data,
  protQ       = PROTEIN_QUANTIFICATION_METHOD,
  method      = NORMALIZATION_METHOD,
  output_path = OUTPUT_PATH
)
```

## Differential expression — DEG and DEP

```{r read-DEG}
source("sources/read_data.R")

DEG_PVAL_THRESHOLD <- 0.05
DE_FC_THRESHOLD    <- 0.5

deg_raw <- purrr::map_dfr(rna_files, function(filepath) {
  species <- tools::file_path_sans_ext(
    stringr::str_extract(basename(filepath), "^[A-Za-z]+_[A-Za-z]+")
  )
  message("  Loading DEG data for: ", species)
  read_deg_data(filepath) |>
    dplyr::mutate(Species = species) |>
    dplyr::select(Species, Protein_id, Treatment, log2FC, fdr, max_mean)
}) |>
  dplyr::filter(!is.na(log2FC), !is.na(fdr), Protein_id %in% common_ids)

# Summarise DEG counts per species and treatment
deg_data <- deg_raw |>
  dplyr::group_by(Species, Treatment) |>
  dplyr::summarise(
    n_de    = sum(fdr < DEG_PVAL_THRESHOLD & abs(log2FC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = dplyr::n(),
    pct_de  = 100 * n_de / n_total,
    de_ids  = list(Protein_id[fdr < DEG_PVAL_THRESHOLD &
                                abs(log2FC) > DE_FC_THRESHOLD &
                                !is.na(fdr)]),
    .groups = "drop"
  )
```

```{r compute-DEP}
source("sources/differential_expression.R")

DEP_PVAL_THRESHOLD <- 0.05

dep_data <- purrr::map_dfr(sort(unique(norm_data$Species)), function(species) {
  message("  Running limma DEP for: ", species)
  sp_data   <- norm_data |>
    dplyr::filter(
      Species   == species,
      Treatment %in% c("Ctrl", STRESS_TREATMENTS_PROT),
      Protein_id %in% common_ids,
      is.finite(value)
    )
  prot_expr <- build_expression_matrix(sp_data)
  run_limma_dep(
    mat            = prot_expr$mat,
    metadata       = prot_expr$metadata,
    species        = species,
    pval_threshold = DEP_PVAL_THRESHOLD,
    fc_threshold   = DE_FC_THRESHOLD
  )
})
```

```{r de-comparison}
source("sources/differential_expression.R")

plots <- plot_de_barplot_all(deg_data, dep_data, OUTPUT_PATH)

combined <- patchwork::wrap_plots(plots, nrow = 1) +
  patchwork::plot_layout(guides = "collect", axes = "collect_y") &
  ggplot2::theme(legend.position = "top")

print(combined)

output_file <- file.path(OUTPUT_PATH, "DEG_DEP_barplot")
save_plot(combined, output_file, width = 16, height = 5)
message("DEG/DEP barplot saved: ", output_file, MSG_FIG_FORMAT)
```

```{r de-correlation}
source("sources/differential_expression.R")

plots <- plot_de_vs_correlation_all(
  deg_data, dep_data, corr_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD, OUTPUT_PATH
)

combined <- patchwork::wrap_plots(plots, nrow = 2) +
  patchwork::plot_layout(guides = "collect", axes = "collect_x") &
  ggplot2::theme(legend.position = "right")

print(combined)

output_file <- file.path(OUTPUT_PATH, "DE_vs_correlation")
save_plot(combined, output_file, width = 12, height = 8)
message("DE vs correlation figure saved: ", output_file, MSG_FIG_FORMAT)
```

---

# Genes with discordantly regulated mRNA and protein levels under stress

```{r zscore-data-generation}
source("sources/discordant_analyses.R")

filtered_protQ <- dplyr::select(protQ_data, Species, Protein_id, Treatment, dplyr::all_of(protQ_meanlog2))

zscore_tpm_df   <- zscore_transform(tpm_log2_data, "mean_log2_TPM")
zscore_protQ_df <- zscore_transform(filtered_protQ, protQ_meanlog2)

discordance_df <- build_discordance_clusters_zscore(
  deg_zscore  = zscore_tpm_df,
  dep_zscore  = zscore_protQ_df,
  z_threshold = 1
)

utils::write.table(
  discordance_df,
  file.path(OUTPUT_PATH, "discordance_clusters.tsv"),
  sep = "\t", row.names = FALSE, quote = FALSE
)

# Discordance summary table (counts + percentages per cluster)
all_proteins <- dplyr::inner_join(
  dplyr::distinct(zscore_tpm_df,   Species, Protein_id),
  dplyr::distinct(zscore_protQ_df, Species, Protein_id),
  by = c("Species", "Protein_id")
)

discordant_proteins <- dplyr::distinct(discordance_df, Species, Protein_id, cluster)

summary_table <- discordant_proteins |>
  dplyr::count(Species, cluster) |>
  dplyr::bind_rows(
    dplyr::anti_join(all_proteins, discordant_proteins, by = c("Species", "Protein_id")) |>
      dplyr::count(Species) |>
      dplyr::mutate(cluster = "Non-discordant")
  ) |>
  dplyr::arrange(Species, cluster) |>
  dplyr::group_by(Species) |>
  dplyr::mutate(total = sum(n), pct = round(100 * n / total, 1)) |>
  dplyr::ungroup()

utils::write.table(
  summary_table,
  file.path(OUTPUT_PATH, "discordance_summary.tsv"),
  sep = "\t", row.names = FALSE, quote = FALSE
)
print(summary_table)
```

```{r zscore-scatter-plot}
source("sources/discordant_analyses.R")

# Save per-species scatter plots
invisible(plot_discordance_scatter_zscore(
  discordance_df, zscore_protQ_df, zscore_tpm_df,
  OUTPUT_PATH, save_per_species = TRUE
))

# Combined scatter with shared ±5 z-score limits
plots <- plot_discordance_scatter_zscore(
  discordance_df, zscore_protQ_df, zscore_tpm_df,
  OUTPUT_PATH, zlim = c(-5, 5)
)

legend         <- cowplot::get_legend(plots[[1]] + ggplot2::theme(legend.position = "top"))
plots_no_legend <- purrr::map(plots, ~ .x + ggplot2::theme(legend.position = "none"))

combined_scatter <- cowplot::plot_grid(
  legend,
  cowplot::plot_grid(plotlist = plots_no_legend, nrow = 1),
  ncol = 1, rel_heights = c(0.1, 0.9)
)

print(combined_scatter)

output_file <- file.path(OUTPUT_PATH, "discordance_scatter")
save_plot(combined_scatter, output_file, width = 12, height = 5)
message("Discordance scatter saved: ", output_file, MSG_FIG_FORMAT)
```

```{r zscore-profiles}
source("sources/discordant_analyses.R")

# Per-species profile plots
purrr::walk(sort(unique(discordance_df$Species)), function(species) {
  p <- plot_discordance_profiles(
    discordance_df, OUTPUT_PATH,
    rna_col        = "rna_z",
    prot_col       = "prot_z",
    y_lab          = "Z-score",
    species_filter = species,
    cluster_spacing = 0.1
  )
  output_file <- file.path(OUTPUT_PATH, species, "discordance_profiles")
  save_plot(p, output_file, width = 12, height = 4)
  message("Profiles saved for: ", species, MSG_FIG_FORMAT)
})

# Combined three-species profile plot
combined_profiles <- plot_discordance_profiles(
  discordance_df, OUTPUT_PATH,
  rna_col         = "rna_z",
  prot_col        = "prot_z",
  y_lab           = "Z-score",
  cluster_spacing = 0.1,
  species_spacing = 0.4
)

output_file <- file.path(OUTPUT_PATH, "discordance_profiles")
save_plot(combined_profiles, output_file, width = 12, height = 8)
message("Combined profiles saved: ", output_file, MSG_FIG_FORMAT)
print(combined_profiles)
```

---

# mRNA-protein correlation in stress-specific stimulon genes

```{r stimulon-data-generation}
source("sources/read_data.R")

stimulon_files <- list.files(STIMULON, pattern = "\\.xlsx$", full.names = TRUE)

# Abbreviation → full species name map for stimulon files
species_stimulon2tpm_map <- c(
  "SE" = "Salmonella_enterica",
  "SA" = "Staphylococcus_aureus",
  "YP" = "Yersinia_pseudotuberculosis"
)

stimulon_df <- purrr::map_dfr(stimulon_files, function(file) {
  extract   <- stringr::str_extract(basename(file), "([A-Za-z]+)_stimulon_([A-Za-z]+)", group = c(1, 2))
  species   <- species_stimulon2tpm_map[extract[[1]]]
  treatment <- extract[[2]]
  tibble::tibble(
    Species    = species,
    Treatment  = treatment,
    Protein_id = read_stimulon_vec(file)
  )
})
```

## Stimulon gene correlation vs all-gene correlation

```{r stimulon-correlation}
source("sources/essential_analyses.R")

stimulon_tpm_df   <- dplyr::semi_join(tpm_log2_data, stimulon_df,  by = c("Species", "Protein_id", "Treatment")) |>
  dplyr::filter(is.finite(mean_log2_TPM))
stimulon_protQ_df <- dplyr::semi_join(protQ_data,    stimulon_df, by = c("Species", "Protein_id", "Treatment")) |>
  dplyr::filter(is.finite(.data[[protQ_meanlog2]]))

corr_stimulon_rna_prot <- calculate_correlation(stimulon_tpm_df, stimulon_protQ_df, protQ_meanlog2)

plots <- plot_subset_vs_all_correlation(
  corr_rna_prot, corr_stimulon_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD, "stimulon", OUTPUT_PATH
)

guide_plot <- ggplot2::ggplot(
    data.frame(t = names(TREATMENT_COLOR)), 
    ggplot2::aes(x = 1, y = 1, color = t)
  ) +
  ggplot2::geom_point() +
  ggplot2::scale_color_manual(
    values = TREATMENT_COLOR, 
    limits = names(TREATMENT_COLOR), 
    drop = FALSE,
    name = "Treatment"
  ) +
  # THIS PART FORCES THE MULTI-LINE OR HORIZONTAL LAYOUT
  ggplot2::guides(
    color = ggplot2::guide_legend(
      nrow = 1,           # Change to 2 or 3 if you have many treatments
      title.position = "top",
      label.position = "bottom"
    )
  ) +
  theme_publication() +
  ggplot2::theme(legend.position = "top")

# 2. Extract the legend
shared_legend <- cowplot::get_legend(guide_plot)

# 3. Assemble
# Increase rel_heights for the first element if the legend still looks cut off
combined <- cowplot::plot_grid(
  shared_legend, 
  cowplot::plot_grid(plotlist = plots, ncol = 3), 
  ncol = 1, 
  rel_heights = c(0.15, 0.85) # Give the legend 15% of the total height
)

print(combined)

output_file <- file.path(OUTPUT_PATH, "stimulon_vs_all_correlation")
save_plot(combined, output_file, width = 12, height = 6)
message("Stimulon vs all correlation figure saved: ", output_file, MSG_FIG_FORMAT)
```

## ORA of osmotic stress-specific mRNA–protein associations

```{r ora-analysis}
source("sources/ORA_analysis.R")

ora_output <- file.path(OUTPUT_PATH, "ORA")
organisms  <- c("SE", "SA", "YP")
dir.create(ora_output, recursive = TRUE, showWarnings = FALSE)

# 1. Process MOBILE output
col_names <- c(
    "ID_TPM", "ID_iBAQ",
    "FULLdata_coeff", "abs", "presence_in_FULL",
    "LOGO_OSS_coeff", "abs.8", "presence_in_LOGO_OSS",
    "LOGO_CTRL_coeff", "abs.3", "presence_in_LOGO_CTRL"
)

results <- purrr::map(organisms, function(org) {
  connection_file <-paste0("interctlist_ALL_", org, ".xlsx")
  data <- readxl::read_excel(file.path(ORA, connection_file)) |>
    dplyr::select(c(3,5,10,11,12,13,14,15,16,17,18)) |>
    setNames(col_names)
  process_organism(data, org, ora_output)
}) |>
  setNames(organisms)

# 2. Build GMT files
purrr::walk(organisms, function(org) {
  create_gmt(
    term2gene_file  = file.path(ORA, paste0(org, "_TERM2GENE.txt")),
    term2name_file  = file.path(ORA, paste0(org, "_TERM2NAME.txt")),
    output_gmt_file = file.path(ora_output, paste0("Updated_", org, "_geneset.gmt"))
  )
})

# 3. Run ORA per organism
ora_results <- purrr::map(organisms, function(org) {
  gmt_pathways <- clusterProfiler::read.gmt(
    file.path(ora_output, paste0("Updated_", org, "_geneset.gmt"))
  )
  term2name <- utils::read.delim(file.path(ORA, paste0(org, "_TERM2NAME.txt")), header = TRUE)
  run_ora_organism(org, results[[org]], gmt_pathways, term2name, ora_output)
}) |>
  setNames(organisms)

# 4. Combine across organisms
combined_dependent <- combine_ora_results(purrr::map(ora_results, "dependent"))
combined_specific  <- combine_ora_results(purrr::map(ora_results, "specific"))

utils::write.csv(combined_dependent, file.path(OUTPUT_PATH, "ORA_dependent_combined_CTRL_OSS.csv"), row.names = FALSE)
utils::write.csv(combined_specific,  file.path(OUTPUT_PATH, "ORA_specific_combined_CTRL_OSS.csv"),  row.names = FALSE)

# 5. Dot plots for each condition
purrr::walk(ORA_CONDITIONS, function(cond) {
  combined <- if (grepl("dependent", cond)) combined_dependent else combined_specific
  p <- plot_ora_dotplot(combined, condition = cond, gap = 0.30)
  output_file <- file.path(OUTPUT_PATH, paste0("ORA_", gsub(" ", "_", cond), "_dotplot"))
  save_plot(p, output_file, width = 5, height = 9)
  print(p)
})

message("ORA pipeline complete. Results saved to: ", OUTPUT_PATH)
```

---

# Supplementary tables and figures

```{r supp-config}
SUPP_OUTPUT_PATH <- file.path(OUTPUT_PATH, "Supplementary")
dir.create(SUPP_OUTPUT_PATH, showWarnings = FALSE)
purrr::walk(SPECIES, ~ dir.create(file.path(SUPP_OUTPUT_PATH, .x), showWarnings = FALSE))
```

## Supplementary Table 1 — iBAQ values for all detected proteins

*iBAQ values for all proteins detected by mass spectrometry across three replicates of 10 stress conditions and the unexposed control. Each species is in a separate sheet.*

```{r supp-table-1}
SUPP_T1_HEADER <- paste(
  "Supplementary Table 1. iBAQ values for all genes detected by mass spectrometry",
  "for each replicate of 10 stress conditions and unexposed control conditions.",
  "The values for each species are shown in a separate sheet."
)

wb <- openxlsx::createWorkbook()

protQ_data |>
  dplyr::mutate(
    Treatment_Replicate = paste(Treatment, stringr::str_replace(Replicate, "R", "rep"), sep = "_")
  ) |>
  dplyr::group_by(Species) |>
  dplyr::group_walk(function(sp_df, sp_key) {
    species <- sp_key$Species

    wide_df <- sp_df |>
      dplyr::select(Protein_id, Treatment_Replicate, iBAQ) |>
      tidyr::pivot_wider(
        names_from  = Treatment_Replicate,
        values_from = iBAQ,
        values_fill = NA
      ) |>
      dplyr::arrange(Protein_id)

    col_order  <- c("Protein_id", sort(setdiff(colnames(wide_df), "Protein_id")))
    wide_df    <- dplyr::select(wide_df, dplyr::all_of(col_order))
    sample_cols <- setdiff(col_order, "Protein_id")
    treatments  <- stringr::str_remove(sample_cols, "_rep\\d+$")

    openxlsx::addWorksheet(wb, sheetName = species)

    # Row 1: merged header
    openxlsx::writeData(wb, species, x = SUPP_T1_HEADER, startRow = 1, startCol = 1)
    openxlsx::mergeCells(wb, species, rows = 1, cols = 1:length(col_order))
    openxlsx::addStyle(wb, species, rows = 1, cols = 1,
                       style = openxlsx::createStyle(
                         fontName = "Arial", fontSize = 10,
                         textDecoration = "bold", wrapText = TRUE
                       ))
    openxlsx::setRowHeights(wb, species, rows = 1, heights = 40)

    # Row 2: treatment group headers (merged per treatment)
    treatment_style <- openxlsx::createStyle(
      fontName = "Arial", fontSize = 10, textDecoration = "bold",
      fgFill = "#B8CCE4", halign = "center", border = "Bottom"
    )
    openxlsx::writeData(wb, species, x = "Protein_id", startRow = 2, startCol = 1)
    openxlsx::addStyle(wb, species, style = treatment_style, rows = 2, cols = 1)

    for (trt in unique(treatments)) {
      trt_cols  <- which(treatments == trt) + 1L   # +1 for Protein_id column
      start_col <- min(trt_cols); end_col <- max(trt_cols)
      openxlsx::writeData(wb, species, x = trt, startRow = 2, startCol = start_col)
      if (start_col < end_col) openxlsx::mergeCells(wb, species, rows = 2, cols = start_col:end_col)
      openxlsx::addStyle(wb, species, style = treatment_style,
                         rows = 2, cols = start_col:end_col, gridExpand = TRUE)
    }

    # Row 3: sample column headers
    sample_style <- openxlsx::createStyle(
      fontName = "Arial", fontSize = 10, textDecoration = "bold",
      fgFill = "#D9E1F2", halign = "center", border = "Bottom"
    )
    openxlsx::writeData(wb, species, x = as.data.frame(t(col_order)),
                        startRow = 3, startCol = 1, colNames = FALSE)
    openxlsx::addStyle(wb, species, style = sample_style,
                       rows = 3, cols = 1:length(col_order), gridExpand = TRUE)

    # Row 4+: data
    openxlsx::writeData(wb, species, x = wide_df, startRow = 4, startCol = 1, colNames = FALSE)
    openxlsx::addStyle(wb, species,
                       style = openxlsx::createStyle(fontName = "Arial", fontSize = 10),
                       rows = 4:(nrow(wide_df) + 3), cols = 1:ncol(wide_df), gridExpand = TRUE)
    openxlsx::setColWidths(wb, species, cols = 1:ncol(wide_df), widths = "auto")

    message("Sheet added: ", species, " (", nrow(wide_df), " proteins × ", ncol(wide_df) - 1, " samples)")
  })

output_file <- file.path(SUPP_OUTPUT_PATH, "Supplementary_Table_1.xlsx")
openxlsx::saveWorkbook(wb, output_file, overwrite = TRUE)
message("Supplementary Table 1 saved: ", output_file)
```

## Pairwise replicate correlations (iBAQ, log2)

Pearson correlation coefficient (*R*) between log2-iBAQ values for each pair of the three replicates per condition.

```{r pairwise-comparison, fig.height=24}
source("sources/protein_quantification.R")

n_inf <- sum(is.infinite(protQ_data$iBAQ_log2))
if (n_inf > 0) {
  message(sprintf("Note: %d rows have -Inf iBAQ_log2 (zero iBAQ); excluded per scatter panel.", n_inf))
}

plot2species <- list()
purrr::walk(sort(unique(protQ_data$Species)), function(sp) {
  message("Pairwise replicate comparison for: ", sp)
  p <- plot_replicate_correlations(
    df           = dplyr::filter(protQ_data, Species == sp),
    species_name = sp
  )
  plot2species[[sp]] <<- p
  output_file <- file.path(SUPP_OUTPUT_PATH, sp, "replicate_correlation")
  save_plot(p, output_file, width = 16, height = 24)
  message("Pairwise comparison saved: ", output_file, MSG_FIG_FORMAT)
})
```

### *Salmonella* Typhimurium

```{r plot-pairwise-senterica, fig.height=24, fig.width=16}}
print(plot2species$Salmonella_enterica)
```

### *Staphylococcus aureus*

```{r plot-pairwise-saureus, fig.height=24, fig.width=16}
print(plot2species$Staphylococcus_aureus)
```

### *Yersinia pseudotuberculosis*

```{r plot-pairwise-ypseudotuberculosis, fig.height=24}
print(plot2species$Yersinia_pseudotuberculosis)
```

## Essential vs non-essential gene distributions per treatment

### mRNA expression

```{r essential-per-condition-rna}
source("sources/essential_analyses.R")

plot <- plot_essential_distribution_per_treatment(
  all_df       = tpm_log2_data,
  essential_df = essential_df,
  log2_col     = "mean_log2_TPM",
  x_lab        = "TPM",
  output_path  = SUPP_OUTPUT_PATH
)
print(plot)
save_plot(
  plot, file.path(SUPP_OUTPUT_PATH, "TPM_per_treatment_essential_distribution"),
  width = 9, height = 18
)
```

### Protein expression

```{r essential-per-condition-protein}
source("sources/essential_analyses.R")

plot <- plot_essential_distribution_per_treatment(
  all_df       = protQ_data,
  essential_df = essential_df,
  log2_col     = protQ_meanlog2,
  x_lab        = PROTEIN_QUANTIFICATION_METHOD,
  output_path  = SUPP_OUTPUT_PATH
)
print(plot)
save_plot(
  plot,
  file.path(SUPP_OUTPUT_PATH, paste0(PROTEIN_QUANTIFICATION_METHOD, "_per_treatment_essential_distribution")),
  width = 9, height = 18
)
```

## Normalization method comparison

```{r normalization-comparison}
source("sources/normalization.R")

run_normalization_comparison(
  protQ_data,
  protQ       = PROTEIN_QUANTIFICATION_METHOD,
  output_path = SUPP_OUTPUT_PATH
)
plot_pca_comparison_grid(
  protQ_data  = protQ_data,
  output_path = SUPP_OUTPUT_PATH
)
```

## Supplementary Table 2 — Discordantly regulated genes per cluster

*Z-score expression values and cluster assignment for all discordantly regulated genes. Each species is in a separate sheet.*

```{r supp-table-2}
SUPP_T2_HEADER <- paste(
  "Supplementary Table 2. Differential expression levels of discordantly regulated genes",
  "in gene clusters with different expression patterns.",
  "Gene lists for each species are shown in separate sheets."
)

wb <- openxlsx::createWorkbook()

dplyr::inner_join(
  zscore_tpm_df,   zscore_protQ_df,
  by = c("Species", "Protein_id", "Treatment"), suffix = c(".mRNA", ".Protein")
) |>
  dplyr::inner_join(discordance_df, by = c("Species", "Protein_id", "Treatment")) |>
  dplyr::select(-c("rna_z", "prot_z", "cluster")) |>
  dplyr::mutate(Expression = paste(rna_sig, prot_sig, sep = "-")) |>
  dplyr::select(-c("rna_sig", "prot_sig")) |>
  dplyr::distinct() |>
  dplyr::group_by(Species) |>
  dplyr::group_walk(function(sp_df, sp_key) {
    species <- sp_key$Species
    openxlsx::addWorksheet(wb, sheetName = species)

    openxlsx::writeData(wb, species, x = SUPP_T2_HEADER, startRow = 1, startCol = 1)
    openxlsx::addStyle(wb, species, rows = 1, cols = 1,
                       style = openxlsx::createStyle(
                         fontName = "Arial", fontSize = 10,
                         textDecoration = "bold", wrapText = TRUE
                       ))
    openxlsx::setRowHeights(wb, species, rows = 1, heights = 40)

    # Column headers
    openxlsx::writeData(wb, species, x = as.data.frame(t(colnames(sp_df))),
                        startRow = 2, startCol = 1, colNames = FALSE)
    openxlsx::addStyle(wb, species,
                       style = openxlsx::createStyle(
                         fontName = "Arial", fontSize = 10, textDecoration = "bold",
                         fgFill = "#D9E1F2", halign = "center", border = "Bottom"
                       ),
                       rows = 2, cols = 1:ncol(sp_df), gridExpand = TRUE)

    # Data
    openxlsx::writeData(wb, species, x = sp_df, startRow = 3, startCol = 1, colNames = FALSE)
    openxlsx::addStyle(wb, species,
                       style = openxlsx::createStyle(fontName = "Arial", fontSize = 10),
                       rows = 3:(nrow(sp_df) + 2), cols = 1:ncol(sp_df), gridExpand = TRUE)
    openxlsx::setColWidths(wb, species, cols = 1:ncol(sp_df), widths = "auto")

    message("Sheet added: ", species, " (", nrow(sp_df), " proteins × ", ncol(sp_df) - 1, " columns)")
  })

output_file <- file.path(SUPP_OUTPUT_PATH, "Supplementary_Table_2.xlsx")
openxlsx::saveWorkbook(wb, output_file, overwrite = TRUE)
message("Supplementary Table 2 saved: ", output_file)
```

## Discordant genes show differential expression across conditions

```{r discordant-DE}
source("sources/discordant_analyses.R")

# DEG counts within discordant clusters
deg_discordance_df <- deg_raw |>
  dplyr::inner_join(discordance_df, by = c("Species", "Protein_id", "Treatment")) |>
  dplyr::group_by(Species, Treatment, cluster) |>
  dplyr::summarise(
    n_de    = sum(fdr < DEG_PVAL_THRESHOLD & abs(log2FC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = dplyr::n(),
    pct_de  = 100 * n_de / n_total,
    de_ids  = list(Protein_id[fdr < DEG_PVAL_THRESHOLD &
                                abs(log2FC) > DE_FC_THRESHOLD &
                                !is.na(fdr)]),
    .groups = "drop"
  )

# DEP counts within discordant clusters
dep_raw <- purrr::map_dfr(sort(unique(norm_data$Species)), function(species) {
  sp_data   <- norm_data |>
    dplyr::filter(
      Species   == species,
      Treatment %in% c("Ctrl", STRESS_TREATMENTS_PROT),
      Protein_id %in% common_ids,
      is.finite(value)
    )
  prot_expr <- build_expression_matrix(sp_data)
  extract_dep_raw(prot_expr$mat, prot_expr$metadata, species)
})

dep_discordance_df <- dep_raw |>
  dplyr::inner_join(discordance_df, by = c("Species", "Protein_id", "Treatment")) |>
  dplyr::group_by(Species, Treatment, cluster) |>
  dplyr::summarise(
    n_de    = sum(adj.P.Val < DEP_PVAL_THRESHOLD & abs(logFC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = dplyr::n(),
    pct_de  = 100 * n_de / n_total,
    de_ids  = list(Protein_id[adj.P.Val < DEP_PVAL_THRESHOLD &
                                abs(logFC) > DE_FC_THRESHOLD &
                                !is.na(adj.P.Val)]),
    .groups = "drop"
  )

plots    <- plot_de_barplot_per_cluster(deg_discordance_df, dep_discordance_df, SUPP_OUTPUT_PATH)
base_size <- 32

combined <- patchwork::wrap_plots(plots, nrow = 3) +
  patchwork::plot_annotation(tag_levels = "A") &
  ggplot2::theme(
    axis.title      = ggplot2::element_text(size = base_size + 2),
    axis.text       = ggplot2::element_text(size = base_size),
    legend.position = "right",
    legend.text     = ggplot2::element_text(size = base_size + 2),
    plot.title      = ggplot2::element_text(size = base_size + 4),
    strip.text      = ggplot2::element_text(size = base_size),
    plot.tag        = ggplot2::element_text(size = 60)
  )

print(combined)

output_file <- file.path(SUPP_OUTPUT_PATH, "discordance_DEG_DEP_barplot")
save_plot(combined, output_file, width = 16, height = 20)
message("Discordance DEG/DEP barplot saved: ", output_file, MSG_FIG_FORMAT)
```

## Enrichment analysis of discordant gene clusters

```{r enrichment-loading}
source("sources/discordant_analyses.R")

enrichment_results <- run_enrichment_all(
  discordance_df = discordance_df,
  pval_cutoff    = 0.01,
  qval_cutoff    = 0.2
)

export_enrichment_data <- enrichment_results |>
  tidyr::separate_longer_delim(geneID, delim = "/") |>
  dplyr::select(
    Species, Cluster, Protein_ID, ID, namespace, Description,
    GeneRatio, BgRatio, RichFactor, FoldEnrichment, zScore,
    pvalue, p.adjust, qvalue
  )

output_file <- file.path(SUPP_OUTPUT_PATH, "enrichment_results.tsv")
utils::write.table(export_enrichment_data, output_file, sep = "\t", row.names = FALSE, quote = FALSE)
message("Enrichment results saved: ", output_file)
```

## Discordantly regulated genes are enriched in translation-related processes

```{r enrichment-mapping}
source("sources/discordant_analyses.R")

p <- plot_enrichment_dotplot(
  enrichment_results = enrichment_results,
  fdr_threshold      = 0.05,
  top_n              = NULL,
  output_path        = SUPP_OUTPUT_PATH,
  cluster_spacing    = 0.1
)
print(p)
```

## Supplementary Table 3 — GO enrichment per cluster

*GO enrichment analysis results for discordant gene clusters. Each species is in a separate sheet.*

```{r supp-table-3}
SUPP_T3_HEADER <- paste(
  "Supplementary Table 3. GO enrichment analysis on gene clusters with discordantly",
  "regulated mRNA and protein levels across three species.",
  "Results for each species are shown in separate sheets."
)

col_order <- c(
  "QUALIFIER", "Description", "ID", "GeneRatio", "BgRatio",
  "pvalue", "p.adjust", "qvalue", "geneID", "Count",
  "GeneRatio_calc", "BgRatio_calc"
)

wb <- openxlsx::createWorkbook()

cluster_style <- openxlsx::createStyle(
  fontName = "Arial", fontSize = 10, textDecoration = "bold",
  fgFill = "#fff2cc", halign = "center", border = "Bottom"
)

enrichment_results |>
  dplyr::group_by(Species) |>
  dplyr::group_walk(function(sp_df, sp_key) {
    species <- sp_key$Species
    openxlsx::addWorksheet(wb, sheetName = species)

    openxlsx::writeData(wb, species, x = SUPP_T3_HEADER, startRow = 1, startCol = 1)
    openxlsx::mergeCells(wb, species, rows = 1, cols = 1:length(col_order))
    openxlsx::addStyle(wb, species, rows = 1, cols = 1,
                       style = openxlsx::createStyle(
                         fontName = "Arial", fontSize = 10,
                         textDecoration = "bold", wrapText = TRUE
                       ))
    openxlsx::setRowHeights(wb, species, rows = 1, heights = 40)

    curr_row <- 2L

    sp_df |>
      dplyr::group_by(Cluster) |>
      dplyr::group_walk(function(cl_df, cl_key) {
        cl <- cl_key$Cluster

        # Cluster separator row
        openxlsx::writeData(wb, species,
                            x = paste("Cluster:", CLUSTER_LABELS[[cl]]),
                            startRow = curr_row)
        openxlsx::mergeCells(wb, species, rows = curr_row, cols = 1:length(col_order))
        openxlsx::addStyle(wb, species, style = cluster_style, rows = curr_row, cols = 1)

        # Data rows
        export_df <- cl_df |>
          dplyr::mutate(
            QUALIFIER      = namespace,
            GeneRatio_calc = sapply(
              strsplit(as.character(GeneRatio), "/"),
              function(x) as.numeric(x[1]) / as.numeric(x[2])
            ),
            BgRatio_calc = sapply(
              strsplit(as.character(BgRatio), "/"),
              function(x) as.numeric(x[1]) / as.numeric(x[2])
            )
          ) |>
          dplyr::select(dplyr::all_of(col_order))

        openxlsx::writeData(wb, species, x = export_df,
                            startRow = curr_row + 1L)
        curr_row <<- curr_row + nrow(export_df) + 3L
      })
  })

output_file <- file.path(SUPP_OUTPUT_PATH, "Supplementary_Table_3.xlsx")
openxlsx::saveWorkbook(wb, output_file, overwrite = TRUE)
message("Supplementary Table 3 saved: ", output_file)
```

## Protein and peptide length distributions

```{r prot-length}
source("sources/protein_quantification.R")

fasta_lengths_df <- purrr::map_dfr(fasta_files, function(fp) {
  sp <- tools::file_path_sans_ext(
    stringr::str_extract(basename(fp), "^[A-Za-z]+_[A-Za-z]+")
  )
  fa <- Biostrings::readAAStringSet(fp)
  tibble::tibble(Species = sp, protein_length = Biostrings::width(fa))
})

p1 <- plot_protein_length(fasta_lengths_df)
save_plot(p1, file.path(SUPP_OUTPUT_PATH, "protein_length"), width = 8, height = 8)
message("Protein length plot saved: ", file.path(SUPP_OUTPUT_PATH, "protein_length"), MSG_FIG_FORMAT)

p2 <- plot_peptide_length(all_raw_digests, min_peptide_len = MIN_PEPTIDE_LEN, max_peptide_len = MAX_PEPTIDE_LEN)
save_plot(p2, file.path(SUPP_OUTPUT_PATH, "peptide_length"), width = 8, height = 8)
message("Peptide length plot saved: ", file.path(SUPP_OUTPUT_PATH, "peptide_length"), MSG_FIG_FORMAT)

fig <- p1 / p2 +
  patchwork::plot_annotation(
    theme = theme_publication() +
      ggplot2::theme(
        plot.title    = ggplot2::element_text(size = 13, face = "bold"),
        plot.subtitle = ggplot2::element_text(size = 10, colour = "#5f5e5a", lineheight = 1.4)
      )
  ) &
  ggplot2::theme(plot.background = ggplot2::element_rect(fill = "white", colour = NA))

output_file <- file.path(SUPP_OUTPUT_PATH, "protein_digestion_analyses")
save_plot(fig, output_file, width = 12, height = 16)
message("Protein digestion figure saved: ", output_file, MSG_FIG_FORMAT)
print(fig)
```

---

# Annexes

## Export package list from renv

```{r export-env}
lock <- renv::lockfile_read("renv.lock")

pkg_df <- do.call(rbind, lapply(lock$Packages, function(pkg) {
  data.frame(
    Package          = pkg$Package %||% NA_character_,
    Version          = pkg$Version  %||% NA_character_,
    Source           = pkg$Source   %||% NA_character_,
    stringsAsFactors = FALSE
  )
}))

utils::write.csv(pkg_df, "renv.csv", row.names = FALSE)
message("renv package list exported to: renv.csv")
```

## Generate MOBILE input files

```{r generate-mobile-data}
raw_file <- file.path(OUTPUT_PATH, paste0(PROTEIN_QUANTIFICATION_METHOD, "_data.tsv"))
raw_data <- utils::read.table(raw_file, sep = "\t", header = TRUE)

# Helper: split a grouped data frame into a named list by species
split_by_species <- function(df) {
  df |>
    dplyr::group_by(Species) |>
    dplyr::group_split(.keep = FALSE) |>
    setNames(
      df |>
        dplyr::group_by(Species) |>
        dplyr::group_keys() |>
        dplyr::pull(Species)
    )
}

# Wide format: one column per replicate × treatment sample
full_df <- raw_data |>
  dplyr::select(Protein_id, Species, Replicate, Treatment, dplyr::all_of(PROTEIN_QUANTIFICATION_METHOD)) |>
  dplyr::mutate(
    ID     = Protein_id,
    sample = paste0(Replicate, "_", Treatment)
  ) |>
  dplyr::select(-Protein_id, -Replicate, -Treatment) |>
  tidyr::pivot_wider(names_from = sample, values_from = dplyr::all_of(PROTEIN_QUANTIFICATION_METHOD)) |>
  dplyr::relocate(ID)

# Mean format: one column per treatment (replicate mean)
mean_df <- raw_data |>
  dplyr::select(Protein_id, Species, Treatment, dplyr::all_of(PROTEIN_QUANTIFICATION_METHOD)) |>
  dplyr::mutate(ID = Protein_id) |>
  dplyr::select(-Protein_id) |>
  tidyr::pivot_wider(
    names_from  = Treatment,
    values_from = dplyr::all_of(PROTEIN_QUANTIFICATION_METHOD),
    values_fn   = mean
  ) |>
  dplyr::relocate(ID)

# Species abbreviation map for output file names
sp2abbv <- c(
  "Salmonella_enterica"       = "SE",
  "Staphylococcus_aureus"     = "SA",
  "Yersinia_pseudotuberculosis" = "YP"
)

purrr::iwalk(split_by_species(full_df), function(df, species) {
  out <- file.path(OUTPUT_PATH, paste0("iBAQ_data_", sp2abbv[[species]], ".tsv"))
  utils::write.table(df, out, sep = "\t", row.names = FALSE, quote = FALSE)
  message("MOBILE full data saved: ", out)
})

purrr::iwalk(split_by_species(mean_df), function(df, species) {
  out <- file.path(OUTPUT_PATH, paste0("iBAQ_data_mean_", sp2abbv[[species]], ".tsv"))
  utils::write.table(df, out, sep = "\t", row.names = FALSE, quote = FALSE)
  message("MOBILE mean data saved: ", out)
})
```
