---
title: "Bacterial Stress Responses Lower mRNA-Protein Level Correlations"
author: "Sena Gizem Suer, Yi Yang Lim, Rabia Şen, Cemal Erdem, André Mateus and Kemal Avican"
date: "2026-02-18"
output:
rmdformats::downcute:
code_folding: hide
self_contained: true
thumbnails: false
lightbox: true
pkgdown:
as_is: true
---
# How to use
Make sure R 4.4
1. git clone
2. install renv
  - Trouble with BioCPrarallel Update Rtools to be same as R in renv
  - Exclude option in restore then manual install
3. Download data
```

# Configuration

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.height = 6,
  fig.width = 12
)
```
## Load library
```{r load-environment, results='hide'}
# Load library
library(fs)
library(purrr)
library(patchwork)
library(cowplot)
library(stringr)
library(tidyverse)
library(readxl)
library(showtext)
library(limma)

# Load sources functions
source("sources/utils.R")
```

## Path and variable
```{r configuration, results='hide', error=TRUE}
#' Species Abbreviation Mapping
#' Maps full species names to abbreviations for data parsing
SPECIES <- c(
  "Salmonella_enterica",
  "Staphylococcus_aureus",
  "Yersinia_pseudotuberculosis"
)

SPECIES_ABBV_MAP <- c(
  "Salmonella_enterica" = "SALMT",
  "Staphylococcus_aureus" = "MSSA",
  "Yersinia_pseudotuberculosis" = "YPSTB"
)

#' Treatment List

STRESS_TREATMENTS <- c("As", "Bs", "Li", "Nd", "Ns", "Oss", "Oxs", "Sp", "Tm")
STRESS_TREATMENTS_RNA <- c(STRESS_TREATMENTS, "Mig")
STRESS_TREATMENTS_PROT <- c(STRESS_TREATMENTS, "Hyp")

#' Control and stress treatment conditions
TREATMENTS_LIST_RNA <- c("Ctrl", STRESS_TREATMENTS_RNA)

PLOT_SPECIES_NAMES <- c(
  "Salmonella_enterica" = "S. enterica Typhimurium",
  "Staphylococcus_aureus" = "S. aureus",
  "Yersinia_pseudotuberculosis" = "Y. pseudotuberculosis"
)


#' Exposure Time Label Mapping
#' Maps numeric exposure times (minutes) to display labels
EXPOSURE_TIME_LABELS <- c(
  "10" = "10",
  "20" = "20 min",
  "30" = "30 min",
  "240" = "4 hours",
  "241" = "\u22654 hours"
)

#' Measurement Display Labels
MEASUREMENT_LABELS <- c(
  "corr_intensity" = "Intensity",
  "corr_RI" = "RI",
  "corr_iBAQ" = "iBAQ",
  "corr_iBAQ_mc" = "iBAQ MC"
)

#' Treatment Shape Mapping (ggstar)
TREATMENT_SHAPES <- c(
  "As" = 1,
  "Bs" = 14,
  "Ctrl" = 15,
  "Hyp" = 11,
  "Li" = 7,
  "Nd" = 8,
  "Ns" = 4,
  "Oss" = 22,
  "Oxs" = 21,
  "Sp" = 23,
  "Tm" = 2
)

# Uncomment the quantification you want to use
# PROTEIN_QUANTIFICATION_METHOD <- "RI" # Intensity divided by the number of razor peptide
# PROTEIN_QUANTIFICATION_METHOD <- "Intensity" # Direct intensity from TMT analyses
# TODO boxplot with protein lenght
PROTEIN_QUANTIFICATION_METHOD = "iBAQ" # Total intensity divided by the number of expected peptide after trypsin digestion
# PROTEIN_QUANTIFICATION_METHOD = "iBAQ_MC" Same but miss cleavage parameter filter added

# --- Directories ---------------------------------------------------------------
dirs <- list(
  DATA = file.path("DATA"),
  RNA_DATA = file.path("DATA", "Read_counts"),
  PROTEIN_DATA = file.path("DATA", "Proteome"),
  PROTEIN_FASTA = file.path("DATA", "Fasta_files"), # Uncomment for iBAQ quantification
  ESSENTIAL_GENES = file.path("DATA", "Essential genes"),
  ENRICHMENT = file.path("DATA", "Enrichment"),
  STIMULON = file.path("DATA", "Stimulons"),
  ORA = file.path("DATA", "ORA")
)

walk(dirs, check_dir, return_error = TRUE)

OUTPUT_PATH <- file.path("Analyses_iBAQ")
FIGURE_FORMAT <- c("png", "svg", "pdf") # "eps", "ps", "tex" (pictex), "jpeg", "tiff", "png", "bmp" or "wmf" (windows only)
MSG_FIG_FORMAT <- paste0(".", paste(FIGURE_FORMAT, collapse = "/."))
dir.create(OUTPUT_PATH)
for (species in SPECIES) dir.create(file.path(OUTPUT_PATH, species))

# --- Required files ------------------------------------------------------------
files <- list(
  SAMPLE_MAPPING = file.path(dirs$PROTEIN_DATA, "sample_mapping.xlsx")
)

walk(files, ~if (!file_exists(.x)) stop(.x, " not found"))

# --- Expose to environment (equivalent to your original variables) -------------
list2env(c(dirs, files), envir = .GlobalEnv)

# 1. Search for Arial on the system and load it
# 'family' is what you'll call it in ggplot; 'regular' is the path to the file.
# On most systems, showtext_auto() finds Arial automatically if you add it:
font_add(family = "Arial", regular = "arial.ttf")
windowsFonts(
  mono = windowsFont("Courier New"),
  Arial = windowsFont("Arial")
)
# 2. Extremely important: This activates showtext for all new plots
showtext_auto()
```

# Results
## Data loading and processing
```{r load-tpm, results='hide'}
source("sources/read_data.R")
# Load RNA tpm data
rna_files <- list.files(RNA_DATA, pattern = "\\.xlsx$", full.names = TRUE)

tpm_data <- map_dfr(rna_files, function (rna_file) {
  species <- get_species_tpm(rna_file)
  abbv <- SPECIES_ABBV_MAP[species]
  read_tpm(rna_file, abbv) %>%
    mutate(
      Treatment = ifelse(Treatment == "Mig", "Hyp", Treatment),
      Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE)
    ) %>%
    filter(!is.na(Protein_id), !is.na(TPM)) %>%
    mutate(Species = species)
})
filtered_tpm <- tpm_data %>%
  group_by(Protein_id) %>%
  mutate(TPM_sum = sum(TPM)) %>%
  filter(TPM_sum == 0)

message(
  "Filtered ", nrow(filtered_tpm), "gene with TPM = 0 for all treatment & Replicate"
)

tpm_data <- tpm_data %>% filter_out(Protein_id %in% filtered_tpm$Protein_id)
```

```{r load-prot, results='hide'}
source("sources/read_data.R")
# Load proteomic data
sample_mapping <- read_excel(SAMPLE_MAPPING, col_names = TRUE) %>%
  mutate(Replicate = paste0("R", Replicate))
protein_files <- list.files(
  PROTEIN_DATA,
  pattern = "\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)

filter_log <- list()

filtered_data <- map_dfr(protein_files, function (protein_file) {
  species <- sub(".*/(\\w+)_R\\d+\\.tsv$", "\\1", protein_file)
  replicate <- sub(".*/\\w+_(R\\d+)\\.tsv$", "\\1", protein_file)

  tmt_raw <- read_tmt(protein_file)
  filtered_raw <- filter_tmt(tmt_raw, species, replicate)

  filter_log[[species]][[replicate]] <<- filtered_raw$filter_log

  filtered_raw$filtered_tmt %>%
    mutate(Species = species, Replicate = replicate)
}) %>%
  left_join(sample_mapping, by = c("Species", "Replicate"), relationship = "many-to-many")

# TODO put this in a function
# Build and write filter log
filter_log_df <- map_dfr(filter_log, ~bind_rows(.x)) %>%
  mutate(across(-c("species", "replicate"), as.integer))

message("TMT filtering summary:\n")

write.table(
  filter_log_df,
  file.path(OUTPUT_PATH, "filter_log.tsv"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

# --- Filter for complete proteins ---
completeness_result <- filter_complete_proteins(filtered_data)
tmt_data <- completeness_result$filtered_data
removed_proteins <- completeness_result$removed_proteins

# Write removed proteins table for manual review
if (nrow(removed_proteins) > 0) {
  write.table(
    removed_proteins %>% arrange(Species, Protein_id),
    file.path(OUTPUT_PATH, "removed_proteins_incomplete.tsv"),
    sep = "\t", row.names = FALSE, quote = FALSE
  )
  message("Removed proteins written to: removed_proteins_incomplete.tsv")
}
```

# Trancriptomic and Proteomic analysis of 3 human pathogens under 10 relevant stress conditions

```{r load-annotation, results='hide'}
annotation_files <- list.files(
  PROTEIN_DATA,
  pattern = "^annotation\\.xlsx$",
  recursive = TRUE,
  full.names = TRUE
)

annotation_data <- map_dfr(annotation_files, function (annotation_file) {
  species <- path_file(path_dir(annotation_file))
  read_excel(
    annotation_file,
    col_names = TRUE, na = "N/A"
  ) %>%
    select(locus_tag, old_locus_tag) %>%
    mutate(
      Protein_id = if_else(
        is.na(locus_tag),
        old_locus_tag,
        locus_tag
      ),
      Protein_id = sub("^pi", "PI", Protein_id, ignore.case = FALSE),
      Species = species
    )
})
```

## Percentages of CDSs detected by transcriptomics and proteomics for the three species
```{r cds-distribution}
source("sources/cds_distribution.R")

detected_cds_df <- get_detected_cds(annotation_data, tpm_data, tmt_data)
detected_cds_file <- file.path(OUTPUT_PATH, "Detected_CDS.tsv")
print(detected_cds_df)
write.table(detected_cds_df, detected_cds_file, sep = "\t")
p <- plot_barplot_detected_cds(
  detected_cds_df,
  position = "overlay"
)

print(p)

output_file <- file.path(OUTPUT_PATH, "CDS_detected")
save_plot(
  plot = p,
  filepath = output_file,
  width = 6,
  height = 9,
)
message(
  "CDS detected for mRNA and prtoein saved: ", output_file, MSG_FIG_FORMAT
)
```

## Distribution of transcript abundance across all conditions in the three species

- All condition: light blue,
- Fraction of proteins detected: yellow,
- Transcripts with no corresponding protein detected: dark blue.

```{r plot-transcript-abundance}
# Shared x_max across all species for visual consistency
source("sources/cds_distribution.R")
tpm_log2_data <- tpm_data %>%
  group_by(Species) %>%
  log2_transform(
    value_col = "TPM",
    group_cols = c("Species", "Protein_id", "Treatment"),
    pseudocount = TRUE
  ) %>%
  filter(is.finite(mean_log2)) %>%
  rename(mean_log2_TPM = mean_log2)

x_max <- tpm_log2_data %>%
  summarise(x_max = ceiling(max(mean_log2_TPM, na.rm = TRUE) / 2.5) * 2.5) %>%
  pull(x_max)

species_list <- unique(tpm_log2_data$Species)

plots <- imap(species_list, function (species, i) {
  message("Plotting TPM detection for: ", species)

  classified <- classify_tpm_groups(tpm_log2_data, tmt_data, species)

  p <- plot_tpm_detection_species(classified, species, x_max = x_max)

  output_sp <- file.path(OUTPUT_PATH, species, "TPM_detection_distribution")
  save_plot(
    plot = p,
    filepath = output_sp,
    width = 12,
    height = 6
  )

  message(
    "CDS distribution for ", species, " saved: ", output_sp, MSG_FIG_FORMAT
  )
  if (i > 1) p <- p + theme(axis.title.y = element_blank())
  p
})

# Combined figure — all species side by side
combined <- wrap_plots(plots, nrow = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")

print(combined)

output_file <- file.path(OUTPUT_PATH, "TPM_detection_distribution")
save_plot(
  plot = combined,
  filepath = file.path(OUTPUT_PATH, "TPM_detection_distribution"),
  width = 12,
  height = 6
)

message("Combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

# mRNA and protein level correlation under stress

## Protein quantification
```{r in-silico-digest, eval = PROTEIN_QUANTIFICATION_METHOD %in% c("iBAQ", "iBAQ_MC")}
# This chunk only runs if PROTEIN_QUANTIFICATION_METHOD is "iBAQ" or "iBAQ_MC"
# For iBAQ quantification
MIN_PEPTIDE_LEN <- 6
MAX_PEPTIDE_LEN <- 30
IBAQ_MISS_CLEAVAGE <- 2
source("sources/protein_quantification.R")

fasta_files <- list.files(PROTEIN_FASTA, pattern = "\\.fa$", full.names = TRUE)
results <- process_tryptic(
  fasta_paths = fasta_files,
  output_path = OUTPUT_PATH,
  min_peptide_len = MIN_PEPTIDE_LEN,
  max_peptide_len = MAX_PEPTIDE_LEN,
  missed_cleavages = ifelse(PROTEIN_QUANTIFICATION_METHOD == "iBAQ", 0, IBAQ_MISS_CLEAVAGE)
)
protein2tryptic <- results$observable_peptides
all_raw_digests <- results$all_raw_digests
```
```{r protein-quantification}
source("sources/protein_quantification.R")

# Protein quantification method — uncomment the desired one:
# PROTEIN_QUANTIFICATION_METHOD <- "RI"       # Intensity divided by number of razor peptides
# PROTEIN_QUANTIFICATION_METHOD <- "Intensity" # Direct intensity from TMT analyses
# PROTEIN_QUANTIFICATION_METHOD <- "iBAQ"        # Total intensity divided by number of expected tryptic peptides
# PROTEIN_QUANTIFICATION_METHOD <- "iBAQ_MC"  # Same as iBAQ but with missed cleavage filter

message("Quantification method: ", PROTEIN_QUANTIFICATION_METHOD)

# Dispatch table: method -> quantification function
quant_fn <- switch(PROTEIN_QUANTIFICATION_METHOD,
                   RI = function (x) RI_quantification(x),
                   Intensity = function (x) Intensity_quantification(x),
                   iBAQ = function (x) ibaq_quantification(x, protein2tryptic, "n_peptides", OUTPUT_PATH),
                   iBAQ_MC = function (x) ibaq_quantification(x, protein2tryptic, "n_peptides", OUTPUT_PATH),
                   stop("Unknown quantification method: ", PROTEIN_QUANTIFICATION_METHOD)
)

quant_data <- quant_fn(tmt_data)
completeness_result <- filter_complete_proteins(
  quant_data, PROTEIN_QUANTIFICATION_METHOD
)
protQ_data <- completeness_result$filtered_data
removed_proteins <- completeness_result$removed_proteins

if (nrow(removed_proteins) > 0) {
  removed_path <- file.path(
    OUTPUT_PATH,
    paste0(PROTEIN_QUANTIFICATION_METHOD, "_removed_proteins_incomplete.tsv")
  )
  write.table(
    removed_proteins %>% arrange(Species, Protein_id),
    removed_path,
    sep = "\t", row.names = FALSE, quote = FALSE
  )
  message(PROTEIN_QUANTIFICATION_METHOD, " removed proteins written to: ", removed_path)
}
protQ_path <- file.path(
  OUTPUT_PATH,
  paste0(PROTEIN_QUANTIFICATION_METHOD, "_data.tsv")
)
write.table(protQ_data, protQ_path, sep = "\t", row.names = FALSE, quote = FALSE)
message(
  PROTEIN_QUANTIFICATION_METHOD,
  " protein quantification data saved: ",
  protQ_path
)
```

## mRNA-protein levels correlation under different stress conditions and control condition

```{r rna-protein-correlation, fig.height = 18}
source("sources/correlation.R")
protQ_meanlog2 <- paste0(PROTEIN_QUANTIFICATION_METHOD, "_meanlog2")
common_ids <- extract_common_ids(tpm_log2_data, protQ_data)

tpm_log2_data <- tpm_log2_data %>% filter(Protein_id %in% common_ids)
protQ_data <- protQ_data %>% filter(Protein_id %in% common_ids)
corr_rna_prot <- calculate_correlation(tpm_log2_data, protQ_data, protQ_meanlog2)
p <- plot_correlation(
  corr_rna_prot, protQ_meanlog2, PROTEIN_QUANTIFICATION_METHOD, OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0("TMP_", PROTEIN_QUANTIFICATION_METHOD, "_correlation_panels")
)
save_plot(
  p,
  output_file,
  width = 12,
  height = 18,
  units = "in"
)
message("Combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

# Essential genes have higher mRNA-protein level correlations in the three species

```{r get-essential-data}
source("sources/essential_analyses.R")
log2_mean_sd_transform <- function (data, value_col, group_cols = c("Species", "Protein_id"), pseudocount = FALSE) {
  mean_col <- paste0("mean_log2_", value_col)
  sd_col <- paste0("sd_log2_", value_col)

  mean_data <- data %>%
    log2_transform(value_col, group_cols = group_cols, pseudocount = pseudocount) %>%
    filter(is.finite(mean_log2)) %>%
    dplyr::rename(!!mean_col := mean_log2)

  sd_data <- data %>%
    log2_mean_sd(value_col, sd_col, group_cols = group_cols, pseudocount = pseudocount)

  left_join(mean_data, sd_data, by = group_cols)
}

essential_files <- list.files(ESSENTIAL_GENES, pattern = "\\.xlsx$", full.names = TRUE)

essential_df <- map_dfr(essential_files, function (file) {
  species <- str_extract(basename(file), "Essential_genes_([A-Za-z]+)", group = 1)
  tibble(
    Species = SPECIES_ESSENTIAL_TO_TPM_MAP[species],
    Protein_id = read_essential_vec(file)
  )
})

tpm_glob_log2mean_data <- tpm_data %>%
  filter(Protein_id %in% common_ids) %>%
  log2_mean_sd_transform("TPM", pseudocount = TRUE)

protQ_glob_log2mean_data <- protQ_data %>%
  filter(Protein_id %in% common_ids) %>%
  log2_mean_sd_transform(PROTEIN_QUANTIFICATION_METHOD, pseudocount = FALSE)

essential_counts_table <- essential_df %>%
  group_by(Species) %>%
  summarise(n_essential_total = n_distinct(Protein_id), .groups = "drop") %>%
  left_join(
    essential_df %>%
      semi_join(tpm_glob_log2mean_data, by = c("Species", "Protein_id")) %>%
      group_by(Species) %>%
      summarise(n_essential_rna = n_distinct(Protein_id), .groups = "drop"),
    by = "Species"
  ) %>%
  left_join(
    essential_df %>%
      semi_join(protQ_glob_log2mean_data, by = c("Species", "Protein_id")) %>%
      group_by(Species) %>%
      summarise(n_essential_prot = n_distinct(Protein_id), .groups = "drop"),
    by = "Species"
  )

print(essential_counts_table)
```

## Mean mRNA expression frequency of essential and non-essential genes

```{r essential-distribution-rna}
source("sources/essential_analyses.R")
p <- plot_essential_distribution(
  all_df = tpm_glob_log2mean_data,
  essential_df = essential_df,
  log2_col = "mean_log2_TPM",
  x_lab = "TPM",
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0("TPM", "_essential_distribution")
)
save_plot(
  p,
  output_file,
  width = 12,
  height = 6,
  units = "in"
)
message("TPM essential distribution combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

## Mean protein expression frequency of essential and non-essential genes
```{r essential-distribution-prot}
source("sources/essential_analyses.R")
p <- plot_essential_distribution(
  all_df = protQ_glob_log2mean_data,
  essential_df = essential_df,
  log2_col = paste0("mean_log2_", PROTEIN_QUANTIFICATION_METHOD),
  x_lab = PROTEIN_QUANTIFICATION_METHOD,
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0(
    PROTEIN_QUANTIFICATION_METHOD,
    "_essential_distribution"
  )
)
save_plot(
  p,
  output_file,
  width = 12,
  height = 6,
  units = "in"
)
message(
  PROTEIN_QUANTIFICATION_METHOD,
  " essential distribution combined figure saved: ",
  output_file,
  MSG_FIG_FORMAT
)
```

## Essential genes and all genes mRNA-protein level correlation

mRNA-protein level correlations were measured with Pearson correlation coefficient (R).
p values were calculated using a Wilcoxon rank sum test.

```{r essential-correlation}
source("sources/essential_analyses.R")
essential_tpm_df <- tpm_log2_data %>%
  semi_join(essential_df, by = c("Species", "Protein_id")) %>%
  filter(is.finite(mean_log2_TPM))
essential_protQ_df <- protQ_data %>%
  semi_join(essential_df, by = c("Species", "Protein_id")) %>%
  filter(is.finite(.data[[protQ_meanlog2]]))

corr_essential_rna_prot <- calculate_correlation(
  essential_tpm_df, essential_protQ_df, protQ_meanlog2
)

p <- plot_subset_vs_all_correlation(
  corr_rna_prot,
  corr_essential_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD,
  "essential",
  OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0(
    PROTEIN_QUANTIFICATION_METHOD,
    "_essential_vs_all_correlation"
  )
)
save_plot(
  p,
  output_file,
  width = 16,
  height = 5,
  units = "in"
)
message(
  "mRNA-Protein correlation all genes vs essential genes combined figure saved: ",
  output_file,
  MSG_FIG_FORMAT
)
```

## Frequency of standard deviation of mRNA expression
```{r essential-sd-rna}
source("sources/essential_analyses.R")
p <- plot_sd_distribution(
  all_df = tpm_glob_log2mean_data,
  essential_df = essential_df,
  sd_col = "sd_log2_TPM",
  x_lab = "TPM",
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0("TPM", "_essential_sd_distribution")
)
save_plot(
  p,
  output_file,
  width = 12,
  height = 6,
  units = "in"
)
message("TPM essential standard deviation distribution combined figure saved: ", output_file, MSG_FIG_FORMAT)
```

## Frequency of standard deviation of protein expression
```{r essential-sd-prot}
source("sources/utils.R")
source("sources/essential_analyses.R")
p <- plot_sd_distribution(
  all_df = protQ_glob_log2mean_data,
  essential_df = essential_df,
  sd_col = paste0("sd_log2_", PROTEIN_QUANTIFICATION_METHOD),
  x_lab = PROTEIN_QUANTIFICATION_METHOD,
  output_path = OUTPUT_PATH
)
print(p)
output_file <- file.path(
  OUTPUT_PATH,
  paste0(
    PROTEIN_QUANTIFICATION_METHOD,
    "_essential_sd_distribution"
  )
)
save_plot(
  p,
  output_file,
  width = 12,
  height = 6,
  units = "in"
)
message(
  PROTEIN_QUANTIFICATION_METHOD,
  " essential standard deviation distribution combined figure saved: ",
  output_file,
  MSG_FIG_FORMAT
)
```

# Stress responses differentially affect mRNA-protein correlation

## mRNA-protein level correlations for conditions exposed stressors

```{r exposure-time}
source("sources/correlation.R")
treatment_to_exposure_time <- c(
  "As" = 10,
  "Bs" = 10,
  "Hyp" = 240,
  "Li" = 10,
  "Nd" = 30,
  "Ns" = 10,
  "Oss" = 10,
  "Oxs" = 10,
  "Sp" = 241, # Not actual time, but used as value >240 -> >4H
  "Tm" = 20
)

print(corr_rna_prot)
exposure_corr_df <- corr_rna_prot %>%
  filter(Treatment != "Ctrl") %>%
  distinct(Species, Treatment, R) %>%
  mutate(
    exposure_time = treatment_to_exposure_time[Treatment],
    exposure_label = factor(
      EXPOSURE_TIME_LABELS[as.character(exposure_time)],
      levels = EXPOSURE_TIME_LABELS
    )
  )
stats_df <- compute_exposure_correlations(exposure_corr_df)
label_pos <- build_label_position(exposure_corr_df, stats_df)

p <- build_exposure_plot(
  df = exposure_corr_df,
  label_pos = label_pos
)

print(p)
output_file <- file.path(OUTPUT_PATH, paste0("correlation_vs_exposure_time"))
save_plot(
  p,
  output_file,
  width = 12,
  height = 5,
  units = "in"
)
message("Correlation vs exposure time saved : ", output_file, MSG_FIG_FORMAT)
```

## Protein normalization

```{r normalization}
source("sources/normalization.R")

NORMALIZATION_METHOD <- "VSN" # Go to supplementary to run normalization comparison

# Single method normalization — returns tidy tibble
norm_data <- run_normalization(
  protQ_data,
  protQ = PROTEIN_QUANTIFICATION_METHOD,
  method = NORMALIZATION_METHOD,
  output_path = OUTPUT_PATH
)
```

## Stress responses differentially affect mRNA-protein correlation

```{r read-DEG}
DEG_PVAL_THRESHOLD <- 0.05
DE_FC_THRESHOLD <- 0.5
source("sources/read_data.R")
deg_raw <- map_dfr(rna_files, function (filepath) {
  species <- tools::file_path_sans_ext(
    str_extract(basename(filepath), "^[A-Za-z]+_[A-Za-z]+")
  )
  message("  Loading DEG data for: ", species)
  deg_data_sp <- read_deg_data(filepath) %>%
    mutate(Species = species) %>%
    dplyr::select(Species, Protein_id, Treatment, log2FC, fdr, max_mean)
}) %>%
  filter(!is.na(log2FC), !is.na(fdr)) %>%
  filter(Protein_id %in% common_ids)

deg_data <- deg_raw %>%
  group_by(Species, Treatment) %>%
  summarise(
    n_de = sum(fdr < DEG_PVAL_THRESHOLD & abs(log2FC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = n(),
    pct_de = 100 * n_de / n_total,
    de_ids = list(Protein_id[fdr < DEG_PVAL_THRESHOLD &
                               abs(log2FC) > DE_FC_THRESHOLD &
                               !is.na(fdr)]),
    .groups = "drop"
  )
```

```{r compute-DEP}
source("sources/differential_expression.R")
DEP_PVAL_THRESHOLD <- 0.05

species_list <- unique(norm_data$Species)

dep_data <- map_dfr(species_list, function (species) {
  message("  Running limma DEP for: ", species)

  sp_data <- norm_data %>%
    filter(
      Species == species,
      Treatment %in% c("Ctrl", STRESS_TREATMENTS_PROT),
      Protein_id %in% common_ids, # add this
      is.finite(value)
    )

  prot_expr <- build_expression_matrix(sp_data)

  run_limma_dep(
    mat = prot_expr$mat,
    metadata = prot_expr$metadata,
    species = species,
    pval_threshold = DEP_PVAL_THRESHOLD,
    fc_threshold = DE_FC_THRESHOLD
  )
})
```

```{r de-comparison}
source("sources/differential_expression.R")
plots <- plot_de_barplot_all(deg_data, dep_data, OUTPUT_PATH)
combined <- wrap_plots(plots, nrow = 1) +
  plot_layout(guides = "collect", axes = "collect_y") &
  theme(legend.position = "top")

print(combined)

output_file <- file.path(OUTPUT_PATH, "DEG_DEP_barplot")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 16, height = 5
)
message("  Combined plot saved: ", output_file, MSG_FIG_FORMAT)
```


```{r de-correlation}
source("sources/differential_expression.R")
plots <- plot_de_vs_correlation_all(
  deg_data,
  dep_data,
  corr_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD,
  OUTPUT_PATH
)
combined <- wrap_plots(plots, nrow = 2) +
  plot_layout(guides = "collect", axes = "collect_x") &
  theme(legend.position = "right")

print(combined)
output_file <- file.path(OUTPUT_PATH, "DE_vs_correlation")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 12, height = 8
)
```

# Identification of genes whose mRNA-protein levels are discordantly regulated under stress conditions

```{r zscore-data-generation}
source("sources/discordant_analyses.R")
filtered_protQ <- protQ_data %>%
  select(Species, Protein_id, Treatment, all_of(protQ_meanlog2))

zscore_tpm_df <- zscore_transform(tpm_log2_data, "mean_log2_TPM")
zscore_protQ_df <- zscore_transform(
  filtered_protQ,
  protQ_meanlog2
)

discordance_df <- build_discordance_clusters_zscore(
  deg_zscore = zscore_tpm_df,
  dep_zscore = zscore_protQ_df,
  z_threshold = 1
)

write.table(discordance_df,
            file.path(OUTPUT_PATH, "discordance_clusters.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

# Summary table
all_proteins <- inner_join(zscore_tpm_df %>% distinct(Species, Protein_id),
                           zscore_protQ_df %>% distinct(Species, Protein_id),
                           by = c("Species", "Protein_id"))

discordant_proteins <- discordance_df %>% distinct(Species, Protein_id, cluster)

summary_table <- discordant_proteins %>%
  count(Species, cluster) %>%
  bind_rows(
    all_proteins %>%
      anti_join(discordant_proteins, by = c("Species", "Protein_id")) %>%
      count(Species) %>%
      mutate(cluster = "Non-discordant")
  ) %>%
  arrange(Species, cluster) %>%
  group_by(Species) %>%
  mutate(total = sum(n), pct = round(100 * n / total, 1)) %>%
  ungroup()

write.table(summary_table, file.path(OUTPUT_PATH, "discordance_summary.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
print(summary_table)
```

```{r zscore-scatter-plot}
source("sources/discordant_analyses.R")
invisible(plot_discordance_scatter_zscore(
  discordance_df, zscore_protQ_df, zscore_tpm_df,
  OUTPUT_PATH, save_per_species = TRUE
))

plots <- plot_discordance_scatter_zscore(
  discordance_df, zscore_protQ_df, zscore_tpm_df,
  OUTPUT_PATH, zlim = c(-5, 5)
)

legend <- get_legend(
  plots[[1]] + theme(legend.position = "top")
)

plots_no_legend <- map(plots, ~.x + theme(legend.position = "none"))

combined_scatter <- plot_grid(
  legend,
  plot_grid(plotlist = plots_no_legend, nrow = 1),
  ncol = 1,
  rel_heights = c(0.1, 0.9)
)

print(combined_scatter)
save_plot(combined_scatter, file.path(OUTPUT_PATH, "discordance_scatter"), width = 12, height = 5)
```

```{r zscore-profiles}
source("sources/discordant_analyses.R")
species_list <- unique(discordance_df$Species)

invisible(walk(species_list, function (species) {
  p <- plot_discordance_profiles(
    discordance_df,
    OUTPUT_PATH,
    species_filter = species,
    cluster_spacing = 0.1
  )
  save_plot(
    p,
    file.path(OUTPUT_PATH, species, "discordance_profiles"),
    width = 12, height = 4
  )
  message("Profiles saved for: ", species, MSG_FIG_FORMAT)
}))

combined_profiles <- plot_discordance_profiles(
  discordance_df, OUTPUT_PATH,
  cluster_spacing = 0.1,
  species_spacing = 0.4   # more space between species rows
)
save_plot(combined_profiles,
          file.path(OUTPUT_PATH, "discordance_profiles"),
          width = 12, height = 8)
print(combined_profiles)
```
# mRNA-protein level stress-specific stimulon genes under stress[

```{r stimulon-data-generation}
source("sources/read_data.R")
stimulon_files <- list.files(STIMULON, pattern = "\\.xlsx$", full.names = TRUE)
species_stimulon2tpm_map <- c(
  "SE" = "Salmonella_enterica",
  "SA" = "Staphylococcus_aureus",
  "YP" = "Yersinia_pseudotuberculosis"
)
stimulon_df <- map_dfr(stimulon_files, function (file) {
  extract <- str_extract(basename(file), "([A-Za-z]+)_stimulon_([A-Za-z]+)", group = c(1, 2))
  species <- species_stimulon2tpm_map[extract[[1]]]
  treatment <- extract[[2]]
  tibble(
    Species = species,
    Treatment = treatment,
    Protein_id = read_stimulon_vec(file)
  )
})
```
## Stress specific stimulon genes and all genes mRNA-protein level correlation

```{r stimulon-correlation}
source("sources/essential_analyses.R")
stimulon_tpm_df <- tpm_log2_data %>%
  semi_join(stimulon_df, by = c("Species", "Protein_id", "Treatment")) %>%
  filter(is.finite(mean_log2_TPM))
stimulon_protQ_df <- protQ_data %>%
  semi_join(stimulon_df, by = c("Species", "Protein_id", "Treatment")) %>%
  filter(is.finite(.data[[protQ_meanlog2]]))

corr_stimulon_rna_prot <- calculate_correlation(
  stimulon_tpm_df, stimulon_protQ_df, protQ_meanlog2
)

plots <- plot_subset_vs_all_correlation(
  corr_rna_prot,
  corr_stimulon_rna_prot,
  PROTEIN_QUANTIFICATION_METHOD,
  "stimulon",
  OUTPUT_PATH
)

combined <- wrap_plots(plots, ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")

print(combined)
output_file <- file.path(
  OUTPUT_PATH,
  paste0(
    PROTEIN_QUANTIFICATION_METHOD,
    "_stimulon_vs_all_correlation"
  )
)
save_plot(combined, output_file, width = 12, height = 4)

message(
  "mRNA-Protein correlation all genes vs stimulon genes combined figure saved: ",
  output_file,
  MSG_FIG_FORMAT
)
```
## ORA of osmotic stress-specific mRNA–protein associations across three species.
```{r ora-analysis}
source("sources/ora_analysis.R")
ora_output <- file.path(OUTPUT_PATH, "ORA")
organisms <- c("SE", "SA", "YP")
dir.create(ora_output, recursive = TRUE, showWarnings = FALSE)

# ---- 1. Process MOBILE output -------------------------------------------
results <- map(organisms, function (org) {
  data <- read.csv2(file.path(ORA, paste0("connections_ALL_", org, ".csv")))
  process_organism(data, org, ora_output)
}) %>% setNames(organisms)

# ---- 2. Build GMT files --------------------------------------------------
walk(organisms, function (org) {
  create_gmt(
    term2gene_file = file.path(ORA, paste0(org, "_TERM2GENE.txt")),
    term2name_file = file.path(ORA, paste0(org, "_TERM2NAME.txt")),
    output_gmt_file = file.path(ora_output, paste0("Updated_", org, "_geneset.gmt"))
  )
})

# ---- 3. Run ORA per organism --------------------------------------------
ora_results <- map(organisms, function (org) {
  gmt_pathways <- read.gmt(file.path(ora_output, paste0("Updated_", org, "_geneset.gmt")))
  term2name <- read.delim(file.path(ORA, paste0(org, "_TERM2NAME.txt")), header = TRUE)
  run_ora_organism(org, results[[org]], gmt_pathways, term2name, ora_output)
}) %>% setNames(organisms)

# ---- 4. Combine results -------------------------------------------------
combined_dependent <- combine_ora_results(map(ora_results, "dependent"))
combined_specific <- combine_ora_results(map(ora_results, "specific"))

write.csv(combined_dependent,
          file.path(OUTPUT_PATH, "ORA_dependent_combined_CTRL_OSS.csv"),
          row.names = FALSE)
write.csv(combined_specific,
          file.path(OUTPUT_PATH, "ORA_specific_combined_CTRL_OSS.csv"),
          row.names = FALSE)

# ---- 5. Plot dotplots ---------------------------------------------------
walk(ORA_CONDITIONS, function (cond) {
  combined <- if (grepl("dependent", cond)) combined_dependent else combined_specific

  p <- tryCatch(
    plot_ora_dotplot(combined, condition = cond, gap = 0.30),
    error = function (e) {
      message("  Skipping dotplot for '", cond, "': ", e$message)
      NULL
    }
  )

  if (!is.null(p)) {
    filename <- file.path(
      OUTPUT_PATH,
      paste0("ORA_", gsub(" ", "_", cond), "_dotplot.png")
    )
    ggsave(filename, p, width = DOTPLOT_WIDTH, height = DOTPLOT_HEIGHT, dpi = 300)
    message("  Saved: ", filename)
    print(p)
  }
})

message("\nORA pipeline complete. Results saved to: ", OUTPUT_PATH)
invisible(list(dependent = combined_dependent, specific = combined_specific))
```

# Supplementary Tables & figures
```{r supp-config}
library(openxlsx)
SUPP_OUTPUT_PATH <- file.path(OUTPUT_PATH, "Supplementary")
dir.create(SUPP_OUTPUT_PATH)
for (species in SPECIES) dir.create(file.path(SUPP_OUTPUT_PATH, species))
```

## Supplementary Table 1. IBAQ values all genes detected my mass spectrometry for each replicates of 10 stress conditions and unexposed control conditions. The values for each species are shown in a separate sheet.

```{r supp-table-1}
header <- "Supplementary Table 1. IBAQ values all genes detected by mass spectrometry for each replicate of 10 stress conditions and unexposed control conditions. The values for each species are shown in a separate sheet."

wb <- createWorkbook()

protQ_data %>%
  mutate(Treatment_Replicate = paste(Treatment, str_replace(Replicate, "R", "rep"), sep = "_")) %>%
  group_by(Species) %>%
  group_walk(function (sp_df, sp_key) {
    species <- sp_key$Species

    wide_df <- sp_df %>%
      dplyr::select(Protein_id, Treatment_Replicate, iBAQ) %>%
      pivot_wider(
        names_from = Treatment_Replicate,
        values_from = iBAQ,
        values_fill = NA
      ) %>%
      arrange(Protein_id)

    col_order <- c("Protein_id", sort(setdiff(colnames(wide_df), "Protein_id")))
    wide_df <- wide_df %>% dplyr::select(all_of(col_order))

    sample_cols <- setdiff(col_order, "Protein_id")
    treatments <- str_remove(sample_cols, "_rep\\d+$")

    addWorksheet(wb, sheetName = species)

    # --- Row 1: Full header text merged across all columns ---------------
    writeData(wb, sheet = species, x = header,
              startRow = 1, startCol = 1)
    mergeCells(wb, sheet = species,
               rows = 1, cols = 1:length(col_order))
    addStyle(wb, sheet = species,
             style = createStyle(
               fontName = "Arial", fontSize = 10,
               textDecoration = "bold", wrapText = TRUE
             ),
             rows = 1, cols = 1)
    setRowHeights(wb, sheet = species, rows = 1, heights = 40)

    # --- Row 2: Treatment group headers (merged per treatment) -----------
    treatment_style <- createStyle(
      fontName = "Arial", fontSize = 10,
      textDecoration = "bold", fgFill = "#B8CCE4",
      halign = "center", border = "Bottom"
    )

    writeData(wb, sheet = species, x = "Protein_id",
              startRow = 2, startCol = 1)
    addStyle(wb, sheet = species, style = treatment_style,
             rows = 2, cols = 1)

    for (trt in unique(treatments)) {
      trt_cols <- which(treatments == trt) + 1  # +1 for Protein_id
      start_col <- min(trt_cols)
      end_col <- max(trt_cols)

      writeData(wb, sheet = species, x = trt,
                startRow = 2, startCol = start_col)
      if (start_col < end_col) {
        mergeCells(wb, sheet = species,
                   rows = 2, cols = start_col:end_col)
      }
      addStyle(wb, sheet = species, style = treatment_style,
               rows = 2, cols = start_col:end_col, gridExpand = TRUE)
    }

    # --- Row 3: Sample column headers (Treatment_Replicate) --------------
    sample_style <- createStyle(
      fontName = "Arial", fontSize = 10,
      textDecoration = "bold", fgFill = "#D9E1F2",
      halign = "center", border = "Bottom"
    )

    writeData(wb, sheet = species, x = as.data.frame(t(col_order)),
              startRow = 3, startCol = 1, colNames = FALSE)
    addStyle(wb, sheet = species, style = sample_style,
             rows = 3, cols = 1:length(col_order), gridExpand = TRUE)

    # --- Row 4+: Data ----------------------------------------------------
    writeData(wb, sheet = species, x = wide_df,
              startRow = 4, startCol = 1, colNames = FALSE)

    addStyle(wb, sheet = species,
             style = createStyle(fontName = "Arial", fontSize = 10),
             rows = 4:(nrow(wide_df) + 3),
             cols = 1:ncol(wide_df), gridExpand = TRUE)

    setColWidths(wb, sheet = species,
                 cols = 1:ncol(wide_df), widths = "auto")

    message("Sheet added: ", species, " (", nrow(wide_df), " proteins x ",
            ncol(wide_df) - 1, " samples)")
  })

output_file <- file.path(SUPP_OUTPUT_PATH, "Supplementary_Table_1.xlsx")
saveWorkbook(wb, output_file, overwrite = TRUE)
message("Excel file saved: ", output_file)
```

## Pairwise comparison of log2-transformed protein expression value (iBAQ) for each of the three replicates per condition.

Pearson correlation coefficient (R) and student-T test

```{r pairwise-comparison, fig.height = 24}
source("sources/protein_quantification.R")
# Guard: warn if log2_iBAQ has unexpected -Inf (zero iBAQ rows)
n_inf <- sum(is.infinite(protQ_data$iBAQ_log2))
if (n_inf > 0)
  message(sprintf(
    "Note: %d rows have -Inf log2_iBAQ (zero iBAQ). ",
    n_inf
  ), "These are excluded per scatter panel.")

species_list <- sort(unique(protQ_data$Species))
message(sprintf("Found %d species: %s",
                length(species_list),
                paste(species_list, collapse = ", ")))

plot2species <- list()
walk(species_list, function (sp) {
  message("Process pairwise comparison for ", sp)
  sp_df <- protQ_data %>% filter(Species == sp)
  p <- plot_replicate_correlations(
    df = sp_df,
    species_name = sp
  )
  plot2species[[sp]] <<- p
  output_file <- file.path(SUPP_OUTPUT_PATH, sp, "replicate_correlation")
  save_plot(
    plot = p,
    filepath = output_file,
    width = 16, height = 24
  )
  message(" Pairwise comparison saved: ", output_file, MSG_FIG_FORMAT)

})
```
### Salmonella enterica

```{r plot-pairwise-senterica}
print(plot2species$Salmonella_enterica)
```
### Staphylococcus aureus
```{r plot-pairwise-saureus}
print(plot2species$Staphylococcus_aureus)
```
### Yersinia pseudotuberculosis
```{r plot-pairwise-tpseudotuberculosis}
print(plot2species$Yersinia_pseudotuberculosis)
```
## Essential vs non-essential per treatment

### Mean mRNA expression frequency of essential and non-essential genes
```{r essential-per-condition-rna}
source("sources/essential_analyses.R")
plot <- plot_essential_distribution_per_treatment(
  all_df = tpm_log2_data,
  essential_df = essential_df,
  log2_col = "mean_log2_TPM",
  x_lab = "TPM",
  output_path = SUPP_OUTPUT_PATH
)

print(plot)

save_plot(
  plot = plot,
  filepath = file.path(SUPP_OUTPUT_PATH, "TPM_per_treatment_essential_distribution"),
  width = 9, height = 18
)
```
### Mean protein expression frequency of essential and non-essential genes
```{r essential-per-condition-protein}
source("sources/essential_analyses.R")
plot <- plot_essential_distribution_per_treatment(
  all_df = protQ_data,
  essential_df = essential_df,
  log2_col = protQ_meanlog2,
  x_lab = PROTEIN_QUANTIFICATION_METHOD,
  output_path = SUPP_OUTPUT_PATH
)

save_plot(
  plot = plot,
  filepath = file.path(SUPP_OUTPUT_PATH, paste0(
    PROTEIN_QUANTIFICATION_METHOD, "_per_treatment_essential_distribution"
  )
  ),
  width = 9, height = 18
)

print(plot)
```
## Normalization
```{r normalization-comparison}
source("sources/normalization.R")
# Method comparison — generates NormalyzerDE reports and PCA plots

run_normalization_comparison(
  protQ_data,
  protQ = PROTEIN_QUANTIFICATION_METHOD,
  output_path = SUPP_OUTPUT_PATH
)
plot_pca_comparison_grid(
  protQ_data = protQ_data,
  output_path = SUPP_OUTPUT_PATH
)
```
## Differential expression levels of discordantly regulated genes in the genes clusters with different expression patterns. Gene list for each species are showng in separate sheet for each species.

```{r supp-table--2}
header <- " Differential expression levels of discordantly regulated genes in the genes clusters with different expression patterns. Gene list for each species are showng in separate sheet for each species."

wb <- createWorkbook()

supp_table <- zscore_tpm_df %>%
  inner_join(zscore_protQ_df, by = c("Species", "Protein_id", "Treatment"), suffix = c(".mRNA", ".Protein")) %>%
  inner_join(discordance_df, by = c("Species", "Protein_id", "Treatment")) %>%
  select(-c("rna_z", "prot_z", "cluster")) %>%
  mutate(Expression = paste(rna_sig, prot_sig, sep = "-")) %>%
  select(-c("rna_sig", "prot_sig")) %>%
  distinct() %>%
  group_by(Species) %>%
  group_walk(function (sp_df, sp_key) {
    species <- sp_key$Species

    addWorksheet(wb, sheetName = species)

    # --- Row 1: Full header text merged across all columns ---------------
    writeData(wb, sheet = species, x = header,
              startRow = 1, startCol = 1)
    addStyle(wb, sheet = species,
             style = createStyle(
               fontName = "Arial", fontSize = 10,
               textDecoration = "bold", wrapText = TRUE
             ),
             rows = 1, cols = 1)
    setRowHeights(wb, sheet = species, rows = 1, heights = 40)

    # --- Row 2: Column headers -------------------------------------------
    writeData(wb, sheet = species, x = as.data.frame(t(colnames(sp_df))),
              startRow = 2, startCol = 1, colNames = FALSE)
    addStyle(wb, sheet = species,
             style = createStyle(
               fontName = "Arial",
               fontSize = 10,
               textDecoration = "bold",
               fgFill = "#D9E1F2",
               halign = "center",
               border = "Bottom"
             ),
             rows = 2, cols = 1:ncol(sp_df), gridExpand = TRUE)

    # --- Row 3+: Data ----------------------------------------------------
    writeData(wb, sheet = species, x = sp_df,
              startRow = 3, startCol = 1, colNames = FALSE)

    addStyle(wb, sheet = species,
             style = createStyle(fontName = "Arial", fontSize = 10),
             rows = 3:(nrow(sp_df) + 2),
             cols = 1:ncol(sp_df), gridExpand = TRUE)

    setColWidths(wb, sheet = species,
                 cols = 1:ncol(sp_df), widths = "auto")

    message("Sheet added: ", species, " (", nrow(sp_df), " proteins x ",
            ncol(sp_df) - 1, " samples)")
  })

output_file <- file.path(SUPP_OUTPUT_PATH, "Supplementary_Table_2.xlsx")
saveWorkbook(wb, output_file, overwrite = TRUE)
message("Excel file saved: ", output_file)
```
## Discordant genes show a differential expression in all conditions
```{r discordant-DE}
source("sources/discordant_analyses.R")
deg_discordance_df <- deg_raw %>%
  inner_join(discordance_df, by = c("Species", "Protein_id", "Treatment")) %>%
  group_by(Species, Treatment, cluster) %>%
  summarise(
    n_de = sum(fdr < DEG_PVAL_THRESHOLD & abs(log2FC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = n(),
    pct_de = 100 * n_de / n_total,
    de_ids = list(Protein_id[fdr < DEG_PVAL_THRESHOLD &
                               abs(log2FC) > DE_FC_THRESHOLD &
                               !is.na(fdr)]),
    .groups = "drop"
  )

dep_raw <- map_dfr(species_list, function (species) {
  sp_data <- norm_data %>%
    filter(Species == species, Treatment %in% c("Ctrl", STRESS_TREATMENTS_PROT),
           Protein_id %in% common_ids, is.finite(value))
  prot_expr <- build_expression_matrix(sp_data)
  extract_dep_raw(prot_expr$mat, prot_expr$metadata, species)
})

dep_discordance_df <- dep_raw %>%
  inner_join(discordance_df, by = c("Species", "Protein_id", "Treatment")) %>%
  group_by(Species, Treatment, cluster) %>%
  summarise(
    n_de = sum(adj.P.Val < DEP_PVAL_THRESHOLD & abs(logFC) > DE_FC_THRESHOLD, na.rm = TRUE),
    n_total = n(),
    pct_de = 100 * n_de / n_total,
    de_ids = list(Protein_id[adj.P.Val < DEP_PVAL_THRESHOLD &
                               abs(logFC) > DE_FC_THRESHOLD &
                               !is.na(adj.P.Val)]),
    .groups = "drop"
  )

plots <- plot_de_barplot_per_cluster(deg_discordance_df, dep_discordance_df, SUPP_OUTPUT_PATH)
base_size = 32
combined <- wrap_plots(plots, nrow = 3) +
  plot_layout() &
  plot_annotation(tag_levels = "A") &
  theme(
    axis.title = element_text(size = base_size + 2),
    axis.text = element_text(size = base_size),
    legend.position = "right",
    legend.text = element_text(size = base_size + 2),
    plot.title = element_text(size = base_size + 4),
    strip.text = element_text(size = base_size),
    plot.tag = element_text(size = 60)
  )
print(combined)

output_file <- file.path(SUPP_OUTPUT_PATH, "discordance_DEG_DEP_barplot")
save_plot(
  plot = combined,
  filepath = output_file,
  width = 16, height = 20
)
```
## Enrichment dotplots
```{r enrichment-loading}
source("sources/discordant_analyses.R")

enrichment_results <- run_enrichment_all(
  discordance_df = discordance_df,
  pval_cutoff = 0.01,
  qval_cutoff = 0.2
)
export_enrichment_data <- enrichment_results %>%
  separate_longer_delim(geneID, delim = "/") %>%
  select(
    Species,
    Cluster,
    Protein_ID,
    ID,
    namespace,
    Description,
    GeneRatio,
    BgRatio,
    RichFactor,
    FoldEnrichment,
    zScore,
    pvalue,
    p.adjust,
    qvalue
  )
output_file <- file.path(SUPP_OUTPUT_PATH, "enrichment_results.tsv")
write.table(
  export_enrichment_data,
  file = output_file,
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

## Discordantly regulated genes are enriched in translation

```{r enrichment-mapping}
source("sources/discordant_analyses.R")
p <- plot_enrichment_dotplot(
  enrichment_results = enrichment_results,
  fdr_threshold = 0.05,
  top_n = NULL, # limit to top 30 terms per cluster
  output_path = SUPP_OUTPUT_PATH,
  cluster_spacing = 0.1
)
print(p)
```

```{r supp-table--3}
header <- "Supplementary Table 3. GO Enrichment analysis on gene clusters with different discordantly regulated mRNA and protein level patterns in three speieces. The analysis results for each species are shown in different sheets."
wb <- createWorkbook()
col_order <- c("QUALIFIER", "Description", "ID", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "qvalue", "geneID", "Count", "GeneRatio_calc", "BgRatio_calc")
# Use group_split to get a list of dataframes, one per species
enrichment_results %>%
  group_by(Species) %>%
  group_walk(function (sp_df, sp_key) {
    species <- sp_key$Species
    addWorksheet(wb, sheetName = species)
    writeData(wb, sheet = species, x = header,
              startRow = 1, startCol = 1)
    mergeCells(wb, sheet = species,
               rows = 1, cols = 1:length(col_order))
    addStyle(wb, sheet = species,
             style = createStyle(
               fontName = "Arial", fontSize = 10,
               textDecoration = "bold", wrapText = TRUE
             ),
             rows = 1, cols = 1)
    setRowHeights(wb, sheet = species, rows = 1, heights = 40)
    cluster_style <- createStyle(
      fontName = "Arial", fontSize = 10,
      textDecoration = "bold", fgFill = "#fff2cc",
      halign = "center", border = "Bottom"
    )
    curr_row <- 2
    sp_df %>%
      group_by(Cluster) %>%
      group_walk(function (cl_df, cl_key) {
        cl <- cl_key$Cluster
        # Write the "Separator"
        writeData(wb, species, x = paste("Cluster: ", CLUSTER_LABELS[[cl]]), startRow = curr_row)
        mergeCells(wb, sheet = species,
                   rows = curr_row, cols = 1:length(col_order))
        addStyle(wb, sheet = species, style = cluster_style,
                 rows = curr_row, cols = 1)
        # Write the data below the separator
        export_df <- cl_df %>%
          mutate(QUALIFIER = namespace,
                 GeneRatio_calc = sapply(strsplit(as.character(GeneRatio), "/"), function (x) { as.numeric(x[1]) / as.numeric(x[2]) }),
                 BgRatio_calc = sapply(strsplit(as.character(BgRatio), "/"), function (x) { as.numeric(x[1]) / as.numeric(x[2]) })
          ) %>%
          select(QUALIFIER, Description, ID, GeneRatio, BgRatio, pvalue, p.adjust, qvalue, geneID, Count, GeneRatio_calc, BgRatio_calc)
        writeData(wb, species, x = export_df, startRow = curr_row + 1)

        # Update row counter (rows + header + separator + gap)
        curr_row <<- curr_row + nrow(export_df) + 3
      })
  })

output_file <- file.path(SUPP_OUTPUT_PATH, "Supplementary_Table_3.xlsx")
saveWorkbook(wb, output_file, overwrite = TRUE)
message("Excel file saved: ", output_file)
```
## Protein and peptide lengths

```{r prot-length}
source("sources/protein_quantification.R")

fasta_lengths_df <- bind_rows(lapply(fasta_files, function (fp) {
  sp <- tools::file_path_sans_ext(
    stringr::str_extract(basename(fp), "^[A-Za-z]+_[A-Za-z]+"))
  fa <- Biostrings::readAAStringSet(fp)
  tibble::tibble(Species = sp, protein_length = Biostrings::width(fa))
}))
p1 <- plot_protein_length(fasta_lengths_df)

output_file <- file.path(SUPP_OUTPUT_PATH, "protein_length")
save_plot(
  plot = p1,
  filepath = output_file,
  width = 8, height = 8
)
message("  Protein lenght plot saved: ", output_file, MSG_FIG_FORMAT)

p2 <- plot_peptide_length(
  all_raw_digests,
  min_peptide_len = MIN_PEPTIDE_LEN,
  max_peptide_len = MAX_PEPTIDE_LEN
)
output_file <- file.path(SUPP_OUTPUT_PATH, "peptide_length")
save_plot(
  plot = p2,
  filepath = output_file,
  width = 8, height = 8
)
message(" Peptide plot saved: ", output_file, MSG_FIG_FORMAT)
fig <- p1 / p2 +
  patchwork::plot_annotation(
    theme = theme_publication() +
      theme(
        plot.title = element_text(size = 13, face = "bold"),
        plot.subtitle = element_text(size = 10, colour = "#5f5e5a",
                                     lineheight = 1.4)
      )
  ) &
  theme(plot.background = element_rect(fill = "white", colour = NA))

output_file <- file.path(SUPP_OUTPUT_PATH, "protein_digestion_analyses")
save_plot(
  plot = fig,
  filepath = output_file,
  width = 12, height = 16
)
message(" Protein digestion plot saved: ", output_file, MSG_FIG_FORMAT)
print(fig)
```

# Annexes

## Export renv

```{r export-env}
# 1. Read the lockfile
lock <- renv::lockfile_read("renv.lock")

# 2. Extract only the simple, non-list fields
clean_list <- lapply(lock$Packages, function (pkg) {
  # We manually pick the fields that are strings, not lists
  data.frame(
    Package = pkg$Package %||% NA,
    Version = pkg$Version %||% NA,
    Source = pkg$Source %||% NA,
    stringsAsFactors = FALSE
  )
})

# 3. Bind them together (this will now work perfectly)
pkg_df <- do.call(rbind, clean_list)

# 4. Export to a friendly CSV
write.csv(pkg_df, "renv.csv", row.names = FALSE)
```
## Generate DF for MOBILE

```{r generate-mobile-data}
raw_file <- file.path(OUTPUT_PATH, paste0(PROTEIN_QUANTIFICATION_METHOD, "_data.tsv"))
raw_data = read.table(
  raw_file,
  sep = "\t",
  header = TRUE
)
full_df <- raw_data %>%
  select(Protein_id, Species, Replicate, Treatment, all_of(PROTEIN_QUANTIFICATION_METHOD)) %>%
  mutate(sample = paste0(Replicate, "_", Treatment)) %>%
  select(-Replicate, -Treatment) %>%
  pivot_wider(names_from = sample, values_from = all_of(PROTEIN_QUANTIFICATION_METHOD)) %>%
  group_by(Protein_id) %>%
  mutate(ID = Protein_id) %>%
  select(-c(Protein_id)) %>%
  relocate(ID)

mean_df <- raw_data %>%
  select(Protein_id, Species, Replicate, Treatment, all_of(PROTEIN_QUANTIFICATION_METHOD)) %>%
  select(-Replicate) %>%
  pivot_wider(names_from = Treatment, values_from = all_of(PROTEIN_QUANTIFICATION_METHOD), values_fn = mean) %>%
  mutate(ID = Protein_id) %>%
  select(-c(Protein_id)) %>%
  relocate(ID)

full_df_list <- full_df %>%
  group_by(Species) %>%
  group_split(.keep = FALSE) %>%
  setNames(
    full_df %>%
      group_by(Species) %>%
      group_keys() %>%
      pull(Species)
  )

mean_df_list <- mean_df %>%
  group_by(Species) %>%
  group_split(.keep = FALSE) %>%
  setNames(
    mean_df %>%
      group_by(Species) %>%
      group_keys() %>%
      pull(Species)
  )

sp2abbv <- list(
  "Salmonella_enterica" = "SE",
  "Staphylococcus_aureus" = "SA",
  "Yersinia_pseudotuberculosis" = "YP"
)

iwalk(full_df_list, function (df, species) {
  output_file <- file.path("Analyses_iBAQ", paste0("iBAQ_data_", sp2abbv[[species]], ".tsv"))
  write.table(
    df,
    file = output_file,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
})

iwalk(mean_df_list, function (df, species) {
  output_file <- file.path("Analyses_iBAQ", paste0("iBAQ_data_mean_", sp2abbv[[species]], ".tsv"))
  write.table(
    df,
    file = output_file,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
})
```
